---
title: Applied Genetic Evaluation - Solution 1
author: Peter von Rohr
date: "2019-04-15"
output:
  bookdown::pdf_document2:
    toc: false
    number_sections: false
    keep_tex: false
header-includes:
 \usepackage{longtable}
 \usepackage{float}
---

```{r setup, include=FALSE}
# knitr::opts_chunk$set(echo = FALSE, results = 'asis', fig.pos = 'H')
knitr::knit_hooks$set(hook_convert_odg = rmddochelper::hook_convert_odg)
```

## Problem 1: Model Selection
We assume that we have a dataset for the response variable `carcass weight` (CW) and for some predictor variables

* sex (`sex`)
* slaughterhouse (`slh`)
* herd (`hrd`)
* age at slaughter (`age`)
* day of month when animal was slaughtered (`day`) and
* humidity (`hum`)

Use a fixed linear effects model and determine which of the predictor variables are important for the response. 

The data is available from `data_bp_w09.csv`. 


### Hint
* Use the function `lm` in R to fit the fixed linear effects model
* Use Mallow $C_p$ statistic and the adjusted coefficient of determination $R_{adj}^2$ as model selection criteria
* Use the backward model selection approach


### Solution
As preparatory step we have to first read the data from the file

```{r}
s_data_file <- "https://charlotte-ngs.github.io/GELASMSS2019/ex/w09/data_bp_w09.csv"
tbl_modsel <- readr::read_csv2(s_data_file)
```

Before we can do any model fits, we have to convert all fixed effects into `factors`. Fixed effects will be

* `sex`
* `slh`
* `hrd`
* `day`

These must be converted into factors. All other predictors are fit as covariables and can stay as numeric types.

```{r}
tbl_modsel$sex <- as.factor(tbl_modsel$sex)
tbl_modsel$slh <- as.factor(tbl_modsel$slh)
tbl_modsel$hrd <- as.factor(tbl_modsel$hrd)
tbl_modsel$day <- as.factor(tbl_modsel$day)
```


The backward model selection approach starts with the full model. 

```{r}
lm_full <- lm(cw ~ sex + slh + hrd + age + day + hum, data = tbl_modsel)
summary(lm_full)
```

```{r}
lm_relevant <- lm(cw ~ sex + slh + hrd + age, data = tbl_modsel)
summary(lm_relevant)
```

