---
title: "Applied Statistical Methods in Animal Sciences"
author: "Peter von Rohr"
date: "2019-02-11"
url: 'https\://github.com/charlotte-ngs/GELASMSS2019'
github-repo: "charlotte-ngs/GELASMSS2019"
site: bookdown::bookdown_site
documentclass: book
bibliography: [bibliography.bib]
link-citations: true
output: bookdown::gitbook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::knit_hooks$set(hook_convert_odg = rmddochelper::hook_convert_odg)
```

```{r mrtminit, include=FALSE}
mrmt <- rmddochelper::MendeleyExportToolR6$new()
mrmt$set_this_rmd_file(ps_this_rmd_file = ifelse(rstudioapi::isAvailable(), 
                                                 rstudioapi::getSourceEditorContext()$path, 
                                                 rprojroot::thisfile()))
```


# Preface {-}
This document contains the course notes for 

__751-7602-00L Applied Statistical Methods in Animal Sciences__. 

This course gives a short introduction to a collection of statistical methods that I believe are relevant for a wide range of topics in Animal Sciences. These methods include 

* Multiple Linear Least Squares Regression (MLLSR)
* Best Linear Unbiased Prediction (BLUP) which is called GBLUP when applied in the context of genomics
* Least Absolute Shrinkage and Selection Operator (LASSO)
* Bayesian Estimation of Unknown Parameters (BEUP)

The above listed collection of statistical methods all happen to be illustrated around the same topic which is called __Genomic Selection__ (GS). GS was introduced in a seminal paper by [@Meuwissen2001]. This very same paper is used as a building block to explain some of the statistical methods (MLLSR and BEUP) used in this course. Furthermore the same publication illustrates that some methods (MLLSR) are not suitable for all different kinds of datasets. 

The time available for this course is just half a semester. This leaves very little time for the introduction of each topic. As a consequence of that each topic can only be presented very superficially and students are expected to work on their own during the exercise hours. Exercises consist of sets of problems related to each topic. Problems are often to be expected to be solved using the R programming language [@RCoreTeam2017]. 

This version of the course is the fourth edition overall and the first time that the course is taught in English. With each additional iteration of the course, improvements are sought to be implemented. Hence any input from the students are greatly appreciated.


## Course Objectives {-}
The students are familiar with the properties of multiple linear regression and they are able to analyse simple data sets using regression methods. The students know why multiple linear regression cannot be used for problems where the number of parameters exceeds the number of observations. One such problem is the prediction of genomic breeding values used in genomic selection. The students know alternative statistical methods that can be applied in situations where the number of parameters is larger than the number of observations. Examples of such methods are BLUP-based approaches, Bayesian procedures and LASSO. The students are able to solve simple exercise problems applying BLUP-based approaches, LASSO and BEUP. The students are expected to use the statistical language and environment R [@RCoreTeam2017].
