---
title: Applied Statistical Methods - Solution 3
author: Peter von Rohr
date: 2019-03-11
output:
  bookdown::pdf_document2:
    toc: false
    number_sections: false
    keep_tex: false
header-includes:
 \usepackage{longtable}
 \usepackage{float}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, results = 'asis', fig.pos = 'H')
knitr::knit_hooks$set(hook_convert_odg = rmddochelper::hook_convert_odg)
```


## Problem 1: Fixed Linear Effects Model {-}
We want to analyse a dataset with genetic information using a fixed linear effects model. The dataset is taken from the course notes and is shown in Table \@ref(tab:dataflemsnpobs). 

We assume that the SNP loci have a purely additive effect on the trait. That means for a SNP locus $L$ the absolute value of the genotypic value of the homozygous genotypes ($L_1L_1$ and $L_2L_2$) is taken to be $a_L$ and the genotypic value of the heterozygous genotype ($L_1L_2$) is taken to be $0$. The fixed linear effects model contains the observation in Table \@ref(tab:dataflemsnpobs) as the response variable, an intercept and the genotypic values of the the genotypes at the two SNP Loci $G$ and $H$ as predictor variables.

For the observation $y_i$ of animal $i$, we can specify the model as

$$y_i = \beta_0 + W_i \cdot a + \epsilon_i$$

where $\beta_0$ is the intercept, $a$ is the vector of additive SNP-effects, $W_i$ is a row vector denoting the SNP-Genotypes and $\epsilon_i$ is the random error term.

```{r dataflemsnpobs, echo=FALSE}
### # fix the number of animals
n_nr_animal <- 20
### # intercept
n_inter_cept <- 500
### # residual standard deviation
n_res_sd <- 12.13
### # vector of genotype value coefficients
vec_geno_value_coeff <- c(-1,0,1)
### # sample genotypes of unlinked SNP randomly
set.seed(436)
### # fix allele frequency of positive allele
n_prob_snps <- .45
### # genotypic values 
vec_geno_val <- c(27.2, 7.3)
n_nr_snp <- length(vec_geno_val)
### # put together the genotypes into a matrix
mat_geno_snp <- matrix(c(sample(vec_geno_value_coeff, n_nr_animal, prob = c((1-n_prob_snps)^2, 
                                                                   2*(1-n_prob_snps)*n_prob_snps, 
                                                                   n_prob_snps^2), 
                       replace = TRUE),
                       sample(vec_geno_value_coeff, n_nr_animal, prob = c(n_prob_snps^2, 
                                                                   2*(1-n_prob_snps)*n_prob_snps, 
                                                                   (1-n_prob_snps)^2), 
                       replace = TRUE)),
                       nrow = n_nr_snp)
mat_obs_y <- n_inter_cept + crossprod(mat_geno_snp, vec_geno_val) + rnorm(n = n_nr_animal, mean = 0, sd = n_res_sd)
### # mapping the -1,0,1 codes to genotypes
geno_code_map <- tibble::tibble(code = c(-1, 0, 1),
                            `SNP G` = c("$G_2G_2$", "$G_1G_2$", "$G_1G_1$"),
                            `SNP H` = c("$H_2H_2$", "$H_1H_2$", "$H_1H_1$"),
                            `Genotypic Value G` = c("$-a_G$", "$0$", "$a_G$"),
                            `Genotypic Value H` = c("$-a_H$", "$0$", "$a_H$"))
geno_code <- tibble::tibble(`Code G` = mat_geno_snp[1,],
                            `Code H` = mat_geno_snp[2,])

### # use dplyr to macht genotypes to codes
suppressPackageStartupMessages( require(dplyr) )
geno_code %>% 
  inner_join(geno_code_map, by = c("Code G" = "code")) %>%
  select(`SNP G`, `Genotypic Value G`) -> geno_snp_g
geno_code %>% 
  inner_join(geno_code_map, by = c("Code H" = "code")) %>%
  select(`SNP H`, `Genotypic Value H`) -> geno_snp_h
geno_snp_all <- bind_cols(geno_snp_g, geno_snp_h)
### # add the data
mat_obs_y_rounded <- round(mat_obs_y, digits = 0)
tbl_obs <- tibble::tibble(Observation = mat_obs_y_rounded[,1])
geno_snp_all %>% bind_cols(tbl_obs) -> tbl_all_data
### # add animal ids
tbl_all_data <- bind_cols(Animal = c(1:n_nr_animal),tbl_all_data)

# tbl_flemsnppbs <- tibble::tibble()
knitr::kable(tbl_all_data,
             booktabs = TRUE,
             longtable = TRUE,
             caption = "Animals With Two SNP Loci Affecting A Quantitative Trait",
             escape = FALSE)
```


### Your Tasks {-}
* Specify the fixed linear effects model in matrix-vector notation by putting the information from the dataset into the model. Use the same parametrization as shown in the course notes where the intercept $\beta_0$ and the vector $a$ are combined into a single parameter vector $b$. The design matrix that links elements in $b$ to observations $y$ is then called $X$. 
* Use the function `rank()` on the matrix $X$ to find out the rank of the design matrix.
* Depending on the rank of $X$ compute an estimate for $b$
* Verify your results using the `lm()` function

### Solution {-}



## Problem 2: 


### Solution

