---
output:
  pdf_document:
    includes:
      in_header: tex/header.tex
    fig_caption: false
output_file: gelss2019_exam_sol.pdf
---

<!-- %\usepackage{fancyhdr} -->

\newcommand{\points}[1]
{\begin{flushright}\textbf{#1}\end{flushright}}

<!-- %\begin{document} -->
<!-- %\SweaveOpts{concordance=TRUE} -->
```{r ChunkOptions, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE, results = 'hide')
#knitr::opts_chunk$set(concordance=TRUE)
knitr::knit_hooks$set(hook_convert_odg = rmddochelper::hook_convert_odg)
```

<!-- <<PointsQ1, echo=FALSE>>= -->
```{r PointsQ1, echo=FALSE}
# Assign Points for Q1
lPointsQ1 <- list(TaskA = 16,
                  TaskB = 6,
                  TaskC = 3,
                  TaskD = 0)
nPointQ1Total <- sum(unlist(lPointsQ1))
```
<!-- @ -->
<!-- <<PointsQ2, echo=FALSE>>= -->
```{r PointsQ2, echo=FALSE}
# Assign Points for Q2
lPointsQ2 <- list(TaskA = 12,
                  TaskB = 8,
                  TaskC = 6,
                  TaskD = 0)
nPointQ2Total <- sum(unlist(lPointsQ2))
```
<!-- @ -->
<!-- <<PointsQ3, echo=FALSE>>= -->
```{r PointsQ3, echo=FALSE}
# Assign Points for Q3
lPointsQ3 <- list(TaskA = 2,
                  TaskB = 11,
                  TaskC = 8,
                  TaskD = 0)
nPointQ3Total <- sum(unlist(lPointsQ3))
```
<!-- @ -->
<!-- <<PointsQ4, echo=FALSE>>= -->
```{r PointsQ4, echo=FALSE}
# Assign Points for Q4
lPointsQ4 <- list(TaskA = 19,
                  TaskB = 20,
                  TaskC = 0,
                  TaskD = 0)
nPointQ4Total <- sum(unlist(lPointsQ4))
```
```{r PointsQ5, echo=FALSE}
# Assign Points for Q4
lPointsQ5 <- list(TaskA = 0,
                  TaskB = 0,
                  TaskC = 0,
                  TaskD = 0)
nPointQ5Total <- sum(unlist(lPointsQ5))
```
<!-- @ -->
<!-- <<PointsTotal, echo=FALSE>>= -->
```{r PointsTotal, echo=FALSE}
nPointOverallTotal <- nPointQ1Total + nPointQ2Total + nPointQ3Total + nPointQ4Total + nPointQ5Total
```
<!-- @ -->


\thispagestyle{empty}

\fcolorbox{white}{white}{
	\centering \parbox[t]{1.0\linewidth}{
		\fontsize{12pt}{20pt}\selectfont % 
		\vspace*{0.5cm} % 

   	\hfill 	Peter von Rohr \\ Institute of Agricultural Sciences\\ D-USYS\\ ETH Zurich

		\vspace*{0.5cm} 
	}
}

\vspace*{2cm}

\fcolorbox{white}{white}{
	\parbox[t]{1.0\linewidth}{
		\centering \fontsize{25pt}{40pt}\selectfont %
		\vspace*{0.2cm}
		 751-6212-00L V \\
     Solutions to Exam    \\
     Applied Genetic Evaluation\\
     SS 2019

		\vspace*{0.7cm} % Space between the end of the title and the bottom of the grey box
	}
}


\vspace*{1cm}

<!-- % Table with Name -->
\begin{tabular}{p{3cm}p{6cm}}
Date:     &  27th May 2019 \\
          &  \\
          &  \\          
Name:     &  \\
          &  \\
          &  \\
Legi-Nr:  & \\
\end{tabular}

<!-- % Table with Points -->

\vspace{5ex}
\begin{center}
\begin{tabular}{|p{3cm}|c|c|}
\hline
Problem  &  Maximum Number of Points  &  Number of Points Reached\\
\hline
1        &  `r nPointQ1Total`  & \\
\hline
2        &  `r nPointQ2Total`  & \\
\hline
3        &  `r nPointQ3Total`  & \\
\hline
4        &  `r nPointQ4Total`  & \\
\hline
Total    &  `r nPointOverallTotal` & \\
\hline
\end{tabular}
\end{center}

\vspace{1cm}

\textit{Questions in German are in italics}

\clearpage
\pagebreak

## Problem 1: Model Selection

\begin{enumerate}
\item[a)] Model selection can be done using \textbf{forward} selection or \textbf{backward} selection. Describe the four steps that are needed in both model selection procedures.

\textit{Modellselektion kann mit \textbf{Forward} Selektion oder \textbf{Backward} Selektion gemacht werden. Beschreiben Sie die vier Schritte, welche es für beide Modellselektionsprozeduren braucht.}
\end{enumerate}
\points{`r lPointsQ1$TaskA`}

### Solution

\begin{tabular}{|l|p{6cm}|p{6cm}|}
\hline
  &  &  \\
Step  &  Forward   &  Backward  \\
  &  &  \\
\hline
  &  &  \\
1  &  Start with smallest model $\mathcal{M}_0$ 
   &  Start with full model $\mathcal{M}_0$ \\
   &  &  \\
\hline
   &  &  \\
2  &  Include predictor reducing RSS the most
   &  Eliminate predictor increasing RSS the least \\
   &  &  \\
\hline
   &  &  \\
3  &  Continue step 2 until all predictors chosen
   &  Continue step 2 until all predictors eliminated  \\
   &  &  \\
\hline
   &  &  \\
4  &  From sequence of submodels $\mathcal{M}_0 \subseteq \mathcal{M}_1 \subseteq \mathcal{M}_2 \subseteq \ldots$ choose model with smallest $C_p$
   &  From sequence of submodels $\mathcal{M}_0 \supseteq \mathcal{M}_1 \supseteq \mathcal{M}_2 \supseteq \ldots$ choose model with smallest $C_p$ \\
   &  &  \\
\hline
\end{tabular}


\clearpage
\pagebreak

\begin{enumerate}
\item[b)] We are given the following data set on body weight, breast circumference and shoulder height. Furthermore, the results of fitting the full regression model and a reduced model with only breast circumference as predictor. Compute the Mallow $C_p$ value for both models and decide which of the two models is better based on $C_p$

\textit{Gegeben ist der folgende Datensatz zu Körpergewicht, Brustumfang und Widerristhöhe. Weiter sind auch die Resultate des Fits des vollen Modells und eines reduzierten Models mit nur Brustumfang als beschreibende Variable gegeben. Berechnen Sie den Mallow $C_p$-Wert für beide Modelle und begründen Sie, welches der beiden Modelle besser ist.}
\end{enumerate}

\points{`r lPointsQ1$TaskB`}

```{r dataregressionP01b, echo=FALSE, results='asis', message=FALSE}
### # define constant paths
s_data_root_online <- "https://charlotte-ngs.github.io/GELASMSS2019"
s_data_root_local <- here::here() 
s_data_dir <- "ex/w03/bw_bc_reg.csv"
### # define data path depending on whether we are online or not
b_online <- FALSE
if (b_online){
  s_data_root <- s_data_root_online
} else {
  s_data_root <- s_data_root_local
}
s_data_path <- file.path(s_data_root, s_data_dir)
### # read data from s_data_path
tbl_reg <- readr::read_csv(file = s_data_path)
### # add random shoulder height
n_sh_mean <- 148
n_sh_sd <- 4.3
vec_sh <- rnorm(nrow(tbl_reg), mean = n_sh_mean, sd = n_sh_sd)
tbl_reg_withsh <- dplyr::bind_cols(tbl_reg, tibble::tibble(`Shoulder Height` = round(vec_sh, 0)))
# remove spaces from colnames
colnames(tbl_reg_withsh) <- c("Animal", "BreastCircumference", "BodyWeight", "ShoulderHeight")
# show the table
knitr::kable(tbl_reg_withsh,
             booktabs = TRUE,
             longtable = TRUE)
```

The results of the full model

\textit{Die Resultate des vollen Modells}

```{r FullModelP01b, echo=FALSE, results='markup'}
lm_bw_bc_sh <- lm(`BodyWeight` ~ `BreastCircumference` + `ShoulderHeight`, data = tbl_reg_withsh)
(sy_bw_bc_sh <- summary(lm_bw_bc_sh))
```

\clearpage
\pagebreak

The results of the reduced model

\textit{Die Resultate des reduzierten Modells}

```{r RedModelP01b, echo=FALSE, results='markup'}
lm_bw_bc <- lm(`BodyWeight` ~ `BreastCircumference`, data = tbl_reg_withsh)
(sy_bw_bc <- summary(lm_bw_bc))
```


### Solution 

Mit `olsrr` erhalten wir

\tiny

```{r SolP01b, echo=TRUE, results='markup'}
olsrr::ols_step_best_subset(lm_bw_bc_sh)
```

\normalsize

Das reduzierte Model erhält einen tieferen $C_p$-Wert und ist somit das bessere Modell.


\clearpage
\pagebreak

\begin{enumerate}
\item[c)] What is the reason why we are doing model selection, why should we not take always the full model? Please explain the underlying phenomenon in your own words.

\textit{Was ist der Grund, dass wir Modellselektion machen? Weshalb können wir nicht einfach immer das volle Modell mit allen erklärenden Variablen zur Anpassung verwenden? Bitte erklären Sie das unterliegende Phänomen in Ihren eigenen Worten.}
\end{enumerate}

\points{`r lPointsQ1$TaskC`}

### Solution 

The problem that makes model selection necessary is called Bias-Variance trade-off. The Bias is reduced, the more predictors we include. But on the other hand, the variance is increased. Hence we have to find the model that balances this trade-off. This is done with a criterion such as the Mallow $C_p$ statistic.



\clearpage
\pagebreak

## Problem 2: Breeding Programs
The breeding goal 2021 (\verb+https://homepage.braunvieh.ch/xml_1/internet/de/application/d2/d9/f464.cfm+) for the `Brown Swiss` cattle breed contains the following information

\textit{Das Zuchtziel 2021 (}\verb+https://homepage.braunvieh.ch/xml_1/internet/de/application/d2/d9/f464.cfm+ \textit{) für das Braunvieh enthält die folgenden Informationen}
\begin{tabular}{|p{6cm}|p{6cm}|}
\hline
Merkmal                                   &  Zielwert für Braunvieh  \\
\hline
Jährlicher Zuchtfortschritt Milch	        &  $+60$ kg \\
\hline
Jährlicher Zuchtfortschritt Eiweiss       &  $+2$ kg  \\
\hline
Milchleistung Ausgewachsene Kuh Talzone  	&  $8500$ kg       \\
\hline
Milchgehalt	Eiweiss                       &  $3.5 \%$          \\
\hline
Laktationspersistenz	                    &  $85 \%$ und höher \\
\hline
Nutzungsdauer (ZWS)	                      &  positiver genetischer Trend \\
\hline
Lebensleistung bei Abgang (Talzone)	      &  $35000$ kg\\
\hline
Eutergesundheit (ZW Zellzahl)	            &  positiver genetischer Trend \\
\hline
ZW Rastzeit	                              &  positiver genetischer Trend\\
\hline
ZW Non Return Rate	                      &  positiver genetischer Trend\\
\hline
Serviceperiode (phänotypisch)	            &  unter 120 Tagen\\
\hline
Rahmen, Becken, Fundament, Euter, Zitzen  &  positiver genetischer Trend\\
\hline
Grösse ausgewachsene Kühe                 &  $140$ - $152$ cm \\
\hline
\end{tabular}


\vspace{10ex}
The aggregate genotype of the Swiss `Edelschein` dam lines has the following weights for the different trait groups `Conformation`, `Production` and `Reproduction`.

\textit{Im Gesamtzuchtwert der Edelschwein Mutterlinie werden die verschiedenen Merkmalsgruppen `Exterieur`, `Produktion` und `Reproduktion` wie folgt gewichtet.}

\textit{Das Zuchtziel 2021 (}\verb+https://homepage.braunvieh.ch/xml_1/internet/de/application/d2/d9/f464.cfm+ \textit{) für das Braunvieh enthält die folgenden Informationen}

\begin{tabular}{|p{6cm}|p{6cm}|}
\hline
Merkmalsgruppe  &  Gewicht im Gesamtzuchtwert \\
\hline
Exterieur       &  $19$  \\
\hline
Produktion      &  $30$  \\
\hline
Reproduktion    &  $51$  \\
\hline
\end{tabular}

\clearpage
\pagebreak

\begin{enumerate}
\item[a)] What are the names of the two different formulation of the breeding goals for `Brown Swiss` cattle and for `Edelschein` pigs? Please complete the following table and in doing so, indicate both an advantage and a disadvantage of both formulations.

\textit{Wie heissen die beiden verschiedenen Arten der Formulierung eines Zuchtziels beim Braunvieh und beim Edelschwein? Bitte füllen Sie die nachfolgende Tabelle aus und geben Sie je einen Vorteil und einen Nachteil der beiden Formulierungen an.}
\end{enumerate}
\points{`r lPointsQ2$TaskA`}



### Solution

\begin{center}
\begin{tabular}{|p{4cm}|p{4cm}|p{4cm}|}
\hline
              &  &  \\
              &  Braunvieh  &  Edelschwein  \\
              &  &  \\
\hline
              &  &  \\
Formulation   &  Political breeding goal with extensive description of mostly phenotypic properties of desired model breeding animal
              &  Scientific breeding goal with aggregate genotype as mathematical form of breeding goal \\
              &  &  \\
\hline
              &  &  \\
Advantage     &  Easy to understand and easy to picture the model breeding animal
              &  Based on solid theory, can be applied for all breeding animals, success can be verified \\
              &  &  \\
\hline
              &  &  \\
Disadvantage  &  Not clear what should be done with over-achiever animals. Goals can be contradictory. Success cannot be verified.
              &  Not easy to understand. Difficult to get clear picture for single breeding animals \\
              &  &  \\
\hline
\end{tabular}
\end{center}

\clearpage
\pagebreak

```{r SetupP02b, echo=FALSE, results='hide'}
n_prop_ani_sel <- .05
n_ag_sd <- 10
n_acc_ng <- .85
n_acc_gs <- .65
n_gi_ng <- 7
n_gi_gs <- 2
tbl_agd_param <- tibble::tibble(Parameter = c("proportion of animals selected",
                                              "accuracy of predicted breeding values",
                                              "additive genetic standard deviation",
                                              "generation interval"),
                                `No Genomic Selection` = c(n_prop_ani_sel,
                                                           n_acc_ng,
                                                           n_ag_sd,
                                                           n_gi_ng),
                                `With Genomic Selection` = c(n_prop_ani_sel,
                                                             n_acc_gs,
                                                             n_ag_sd,
                                                             n_gi_gs))
```

\begin{enumerate}
\item[b)] Genomic Selection has the potential to increase the genetic gain per year significantly. Currently beef cattle breeders start to introduce genomic selection for carcass traits. Let us assume the following selection parameters for the trait average daily gain. Compute the selection response per year for both scenarios with and without genomic selection. Based on the selection response per year which scenario is better? The trait is assumed to be normally distributed. 

\textit{Die genomische Selektion hat das Potential den Zuchtfortschritt pro Jahr signifikant zu steigern. Aktuell beginnen die Fleischrinderzüchter mit der Einführung der genomischen Selektion für Fleischleistungsmerkmale. Für das Merkmal Tageszunahme gehen wir von den folgenden Selektionsparametern aus. Berechnen Sie den Selektionserfolg pro Jahr für die beiden Szenarien mit und ohne genomische Selektion. Welches der beiden Szenarien ist besser basierend auf dem Selektionsfortschritt pro Jahr? Für die Verteilung des Merkmals nehmen wir eine Normalverteilung an.}
\end{enumerate}
\points{`r lPointsQ2$TaskB`}

```{r TableP02b, echo=FALSE, results='asis'}
knitr::kable(tbl_agd_param,
             booktabs = TRUE,
             longtable = TRUE)
```


### Solution

```{r SolP02b, echo=FALSE, results='hide'}
n_val_z <- dnorm(qnorm(n_prop_ani_sel, lower.tail = FALSE))
n_sel_int <- n_val_z / n_prop_ani_sel
n_sel_int_round <- round(n_sel_int, 3)
n_delta_g_ng <- (n_sel_int * n_acc_ng * n_ag_sd) / n_gi_ng
n_delta_g_gs <- (n_sel_int * n_acc_gs * n_ag_sd) / n_gi_gs
```

The proportion animals selected has to be converted into a selection intensity $i$.

$$i = \frac{z}{p} = \frac{`r round(n_val_z, 3)`}{`r n_prop_ani_sel`} = `r n_sel_int_round`$$

The selection response is computed as

$$\Delta G = \frac{i*r_{TI} * \sigma_T}{L}$$

**No Genomic Selection**: 

$$\Delta G_{ng} = \frac{`r n_sel_int_round` * `r n_acc_ng` * `r n_ag_sd`}{`r n_gi_ng`} = `r round(n_delta_g_ng, 3)`$$

**With Genomic Selection**: 

$$\Delta G_{gs} = \frac{`r n_sel_int_round` * `r n_acc_gs` * `r n_ag_sd`}{`r n_gi_gs`} = `r round(n_delta_g_gs, 3)`$$

Based on the selection response per year, the scenario with genomic selection is better. 


\clearpage
\pagebreak

\begin{enumerate}
\item[c)] In cattle and in pigs, the structure of the breeding programs are different. What is the name of the structure of the different breeding programs. Please, specify a reason for the association of the structure of the breeding program to the different species.

\textit{In der Rinderzucht und der Schweinezucht ist die Struktur der Zuchtprogramme verschieden. Wie heissen diese Strukturen der Zuchtprogramme? Geben Sie je einen Grund an weshalb die beiden Spezies die entsprechende Struktur des Zuchtprogramms gewählt haben.}
\end{enumerate}
\points{`r lPointsQ2$TaskC`}


### Solution 

\begin{center}
\begin{tabular}{|p{4cm}|p{4cm}|p{4cm}|}
\hline
                               &          &  \\
                               &  Cattle  &  Pigs  \\
                               &          &  \\
\hline
                               &          &  \\
Structure of breeding program  &  monolitic  &  hierarchical  \\    
                               &          &  \\
\hline
                               &          &  \\
Reason                         &  low reproduction rate, single animal is valuable, sex-limited traits make expensive off-spring tests necessary, hence large herdbooks with many member farms are required
                               &  high reproduction rate, single animal not so valuable. Split between breeding and production can be used to increase efficiency \\  
                               &          &  \\
\hline     
\end{tabular}
\end{center}

\clearpage
\pagebreak


## Problem 3: Variance Components Estimation

```{r SetupDataP03}
set.seed(53445)
# according to google and https://timeforchange.org/are-cows-cause-of-global-warming-meat-methane-CO2
n_min_met_prod_per_year <- 70
n_max_met_prod_per_year <- 120
n_av_met_prod_per_year <- mean(c(n_min_met_prod_per_year,n_max_met_prod_per_year))
n_sd_met_prod_per_year <- (n_max_met_prod_per_year - n_min_met_prod_per_year) / 6
# number of bull
n_nr_bull <- 5
vec_bull_eff <- rnorm(n_nr_bull, sd = n_sd_met_prod_per_year / 4)
# number of daughters per bull
n_nr_dau <- 3
# number of observations
n_nr_met_obs <- n_nr_bull * n_nr_dau
# bull incidence matrix
mat_bull_design <- kronecker(diag(n_nr_bull), matrix(rep(1,n_nr_dau), ncol = 1))

# observations
mat_obs <- n_av_met_prod_per_year + crossprod(t(mat_bull_design), vec_bull_eff) + rnorm(n_nr_met_obs, mean = 0, n_sd_met_prod_per_year)
# data tibble
tbl_met <- tibble::tibble(Offspring = c(1:n_nr_met_obs), 
                          Bull = sort(rep(1:n_nr_bull, n_nr_dau)), 
                          Methane = round(mat_obs[,1], 0))

```


The following data set with average amounts of methane gas emission per year for three offspring per bull are given in the following table.

\textit{Im folgenden Datensatz sind die mittleren Mengen an Methanemissionen pro Jahr für je drei Nachkommen für fünf verschiedene Bullen gegeben.}

```{r TableDataP03, echo=FALSE, results='asis'}
knitr::kable(tbl_met,
             booktabs = TRUE,
             longtable = TRUE)
```


\begin{enumerate}
\item[a)] Given that we want to reduce the amount of methane gas emitted in cattle using the tools of livestock breeding, we first have to do a variance components estimation. Why is it important to have a certain variability in a given trait and which variance component is important when we want to improve a certain trait with livestock breeding tools?

\textit{Wir möchten die Methanemission beim Rind mit den Werkzeugen der Tierzucht senken. Weshalb brauchen wir für die züchterische Bearbeitung eines Merkmals ein gewisse Variabilität im Merkmal und welche Varianzkomponente interessiert uns besonders?}
\end{enumerate}
\points{`r lPointsQ3$TaskA`}

### Solution

One of the main livestock breeding tools is selection. Selection means that we choose from a given population the best individuals to be parents of the next generation. Hence selection can only happen, when a certain level of variation can be observed in a population. Because livestock breeding means improvement of a population at the genetic level, the genetic variance component or in our case the sire variance component is important.

\clearpage
\pagebreak


\begin{enumerate}
\item[b)] What is the linear model that would be used to allows us a separation of the different variance components in the given dataset on methane gas emission?

\textit{Wie sieht das lineare Modell aus, welches eine Aufteilung der Varianz in ihre verschiedenen Komponenten im Methandatensatz erlaubt?}
\end{enumerate}

\points{`r lPointsQ3$TaskB`}


### Solution

The model is

$$y_{ij} = \mu + s_i + e_{ij}$$

\begin{tabular}{lll}
where  &  &  \\
       & $y_{ij}$         &  measurement $j$ of animal $i$    \\
       & $\mu$            &  expected value of $y$            \\
       & $s_i$            &  deviation of $y_{ij}$ from $\mu$ attributed to bull $i$ \\
       & $e_{ij}$  &  measurement error
\end{tabular}

The expected values are: $E\left[ s_i \right] = 0$, $E\left[ e_{ij} \right] = 0$, $E\left[ y_{ij} \right] = \mu$

The variances: $var(s_i) = \sigma_s^2$, $var(e_{ij}) = \sigma_e^2$, $var(y_{ij}) = \sigma_s^2 + \sigma_e^2$

\clearpage
\pagebreak

\begin{enumerate}
\item[c)] Use an analysis of variance (ANOVA) to estimate the variance components given in the above data set.

\textit{Schätzen Sie mit einer Varianzanalyse (ANOVA) die Varianzkomponenten aus dem oben angegebenen Datensatz zu den Methanemissionen.}
\end{enumerate}
\points{`r lPointsQ3$TaskC`}

### Solution

The sires are first converted into factors to prevent R to fit them as covariables.

```{r PrepP03c, echo=TRUE, results='markup'}
tbl_met$Bull <- as.factor(tbl_met$Bull)
```

Use `aov()` to get the ANOVA-table

```{r AovP03c, echo=TRUE, results='markup'}
aov_meth <- aov(Methane ~ Bull, data = tbl_met)
(sy_aov_meth <- summary(aov_meth))
```

From this the estimate of the residual variance corresponds to

```{r Sigmae2, echo=TRUE, results='markup'}
(n_hat_sigma_e2 <- sy_aov_meth[[1]]$`Mean Sq`[2])
```

```{r Sigmas2, echo=TRUE, results='markup'}
(n_hat_sigma_s2 <- (sy_aov_meth[[1]]$`Mean Sq`[1] - sy_aov_meth[[1]]$`Mean Sq`[2]) / n_nr_dau)
```


\clearpage
\pagebreak


## Problem 4: Prediction of Breeding Values

We are using the same dataset as in Problem 3. The results of the variance components estimation were convincing that it is possible to reduce methane gas emission via selection. The next step is to predict breeding values. 

\textit{Wir verwenden nochmals den gleichen Datensatz wie in Aufgabe 3. Die Resultate der Varianzkomponenten waren positiv in dem Sinne, dass eine züchterische Bearbeitung des Merkmals Methanemission möglich erscheint.}

```{r TableDataP04, echo=FALSE, results='asis'}
knitr::kable(tbl_met,
             booktabs = TRUE,
             longtable = TRUE)
```

\begin{enumerate}
\item[a)] Specify a sire model by writing down the model formula, describing the meaning of all model components and indicating the expected values and the variances of all random effects in the model. We are using a general mean $\mu$ as the only fixed effect. We assume that all bulls are unrelated. Input the numeric information from the dataset into the model components

\textit{Spezifizieren Sie für die Zuchtwertschätzung ein Vatermodell. Geben Sie die Modellformel, beschreiben Sie die Bedeutung aller Modellkomponenten und geben Sie für alle zufälligen Komponenten im Modell den Erwartungswert und die Varianz an. Als einziger fixer Effekt wird ein allgemeines Mittel $\mu$ angenommen. Die Väter sind nicht miteinander verwandt. Setzen Sie die numerische Information aus dem Datensatz in die Modellkomponenten ein.}
\end{enumerate}
\points{`r lPointsQ4$TaskA`}


### Solution

The sire model is

$$y = Xb + Zs + e$$

\begin{tabular}{lll}
  where  &  &  \\
         &  $y$  &  vector of length $n$ with observations  \\
         &  $b$  &  general mean as the only fixed effect \\
         &  $X$  &  incidence matrix linking elements in $b$ to observations \\
         &  $s$  &  vector of length $t$ with random sire effects \\
         &  $Z$  &  incidence matrix linking elements in $s$ to observations \\
         &  $e$  &  vector of length $n$ of random error terms
\end{tabular}

Inserting the data into the model

```{r ModelInfoP01a, echo=FALSE, results='asis'}
vec_y_met <- tbl_met$Methane
vec_b_met <- c("\\mu")
mat_X_met <- matrix(c(rep(1,n_nr_met_obs)), ncol = 1)
vec_s_met <- rmddochelper::vecGetVecElem(psBaseElement = "s", pnVecLen = n_nr_bull, psResult = "latex") 
mat_Z_met <- mat_bull_design
vec_e_met <- rmddochelper::vecGetVecElem(psBaseElement = "e", pnVecLen = n_nr_met_obs, psResult = "latex") 
cat("$$\n")
cat("y = \\left[\n")
cat(paste0(rmddochelper::sConvertVectorToLaTexArray(pvec_avector = vec_y_met, pnDigits = 0), collapse = "\n"), "\n")
cat("\\right], \n")
cat("b = \\left[\n")
cat(paste0(rmddochelper::sConvertVectorToLaTexArray(pvec_avector = vec_b_met), collapse = "\n"), "\n")
cat("\\right], \n")
cat("X = \\left[\n")
cat(paste0(rmddochelper::sConvertMatrixToLaTexArray(pmatAMatrix = mat_X_met, pnDigits = 0), collapse = "\n"), "\n")
cat("\\right], \n")
cat("s = \\left[\n")
cat(paste0(rmddochelper::sConvertVectorToLaTexArray(pvec_avector = vec_s_met), collapse = "\n"), "\n")
cat("\\right], \n")
cat("Z = \\left[\n")
cat(paste0(rmddochelper::sConvertMatrixToLaTexArray(pmatAMatrix = mat_Z_met, pnDigits = 0), collapse = "\n"), "\n")
cat("\\right], \n")
cat("e = \\left[\n")
cat(paste0(rmddochelper::sConvertVectorToLaTexArray(pvec_avector = vec_e_met), collapse = "\n"), "\n")
cat("\\right]\n")
cat("$$\n")
```

The expected values of the random components are: $E\left[s\right] = 0$, $E\left[e\right] = 0$ and $E\left[y \right] = Xb = 1_n\mu$

The variances are $var(s) = G = A_s * \sigma_s^2$ where $A_s$ is the numerator relationship matrix for the sires. Because the sires are unrelated $A_s = I$. The variance of the errors is given by $var(e) = R = I * \sigma_e^2$ and $var(y) = ZGZ^T + R$. 


\clearpage
\pagebreak

```{r SetupP04b}
n_sigma_s2 <- 4
n_sigma_e2 <- 72
n_lambda <- n_sigma_e2 / n_sigma_s2
```

\begin{enumerate}
\item[b)] Set up the mixed model equations for the model specified in 4a) to estimate the fixed effect and to predict the sire effects. We assume the sire variance to be $\sigma_s^2 = `r n_sigma_s2`$ and the error variance to be $\sigma_e^2 = `r n_sigma_e2`$. 

\textit{Stellen Sie die Mischmodellgleichungen für das Modell aus 4a) auf und berechnen Sie die Schätzung des fixen Effekts und der Vatereffekte.}
\end{enumerate}
\points{`r lPointsQ4$TaskB`}

### Solution

The mixed model equations using the simplifications from 4a) are given by

$$
\left[
  \begin{array}{lr} 
  X^TX  &  X^TZ  \\
  Z^TX  &  Z^TZ + \lambda * I
  \end{array} 
\right]
\left[
  \begin{array}{c}
  \hat{\mu} \\
  \hat{s}
  \end{array} 
\right]
=
\left[
  \begin{array}{c}
  X^Ty \\
  Z^Ty
  \end{array} 
\right]
$$

where $\lambda = \frac{\sigma_e^2}{\sigma_s^2}$

Inserting the numbers leads to

```{r MmeP04b, echo=FALSE, results='hide'}
mat_xtx <- crossprod(mat_X_met)
mat_xtz <- crossprod(mat_X_met, mat_Z_met)
mat_ztz <- crossprod(mat_Z_met)
mat_ztz_lambda <- mat_ztz + n_lambda * diag(n_nr_bull)
mat_coeff <- rbind(cbind(mat_xtx, mat_xtz), cbind(t(mat_xtz), mat_ztz_lambda))
mat_rhs <- rbind(crossprod(mat_X_met, vec_y_met), crossprod(mat_Z_met, vec_y_met))
vec_unknown <- c("\\hat{\\mu}", "\\hat{s}")
mat_sol <- solve(mat_coeff, mat_rhs)
```

```{r ShowMmeP04b, echo=FALSE, results='asis'}
cat("$$\n")
cat("\\left[")
cat(paste0(rmddochelper::sConvertMatrixToLaTexArray(pmatAMatrix = mat_coeff, pnDigits = 0)))
cat("\\right] \n")
cat("\\left[")
cat(paste0(rmddochelper::sConvertVectorToLaTexArray(pvec_avector = vec_unknown)))
cat("\\right] \n")
cat("= \n")
cat("\\left[")
cat(paste0(rmddochelper::sConvertMatrixToLaTexArray(pmatAMatrix = mat_rhs, pnDigits = 0)))
cat("\\right] \n")
cat("$$\n")
```

Solving for the vector of unknowns leads to 

```{r SolveMmeP04b, echo=FALSE, results='asis'}
cat("$$\n")
cat("\\left[")
cat(paste0(rmddochelper::sConvertVectorToLaTexArray(pvec_avector = vec_unknown)))
cat("\\right] \n")
cat("= \n")
cat("\\left[")
cat(paste0(rmddochelper::sConvertMatrixToLaTexArray(pmatAMatrix = mat_coeff, pnDigits = 0)))
cat("\\right]^{-1} \n")
cat("\\left[")
cat(paste0(rmddochelper::sConvertMatrixToLaTexArray(pmatAMatrix = mat_rhs, pnDigits = 0)))
cat("\\right] \n")
cat("= \n")
cat("\\left[")
cat(paste0(rmddochelper::sConvertMatrixToLaTexArray(pmatAMatrix = mat_sol, pnDigits = 3)))
cat("\\right] \n")
cat("$$\n")
```

