---
output:
  pdf_document:
    includes:
      in_header: tex/header.tex
    fig_caption: false
output_file: asmfs2018_exam_tsk.pdf
---

<!-- %\usepackage{fancyhdr} -->

\newcommand{\points}[1]
{\begin{flushright}\textbf{#1}\end{flushright}}

<!-- %\begin{document} -->
<!-- %\SweaveOpts{concordance=TRUE} -->
```{r ChunkOptions, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE, results = 'hide')
#knitr::opts_chunk$set(concordance=TRUE)
knitr::knit_hooks$set(conv.odg = rmddochelper::odg.graphics.conv.hook)
```

<!-- <<PointsQ1, echo=FALSE>>= -->
```{r PointsQ1, echo=FALSE}
# Assign Points for Q1
lPointsQ1 <- list(TaskA = 0,
                  TaskB = 0,
                  TaskC = 0,
                  TaskD = 0)
nPointQ1Total <- sum(unlist(lPointsQ1))
```
<!-- @ -->
<!-- <<PointsQ2, echo=FALSE>>= -->
```{r PointsQ2, echo=FALSE}
# Assign Points for Q2
lPointsQ2 <- list(TaskA = 0,
                  TaskB = 0,
                  TaskC = 0,
                  TaskD = 0)
nPointQ2Total <- sum(unlist(lPointsQ2))
```
<!-- @ -->
<!-- <<PointsQ3, echo=FALSE>>= -->
```{r PointsQ3, echo=FALSE}
# Assign Points for Q3
lPointsQ3 <- list(TaskA = 0,
                  TaskB = 0,
                  TaskC = 0,
                  TaskD = 0)
nPointQ3Total <- sum(unlist(lPointsQ3))
```
<!-- @ -->
<!-- <<PointsQ4, echo=FALSE>>= -->
```{r PointsQ4, echo=FALSE}
# Assign Points for Q4
lPointsQ4 <- list(TaskA = 0,
                  TaskB = 0,
                  TaskC = 0,
                  TaskD = 0)
nPointQ4Total <- sum(unlist(lPointsQ4))
```
```{r PointsQ5, echo=FALSE}
# Assign Points for Q4
lPointsQ5 <- list(TaskA = 0,
                  TaskB = 0,
                  TaskC = 0,
                  TaskD = 0)
nPointQ5Total <- sum(unlist(lPointsQ5))
```
<!-- @ -->
<!-- <<PointsTotal, echo=FALSE>>= -->
```{r PointsTotal, echo=FALSE}
nPointOverallTotal <- nPointQ1Total + nPointQ2Total + nPointQ3Total + nPointQ4Total + nPointQ5Total
```
<!-- @ -->


\thispagestyle{empty}

\fcolorbox{white}{white}{
	\centering \parbox[t]{1.0\linewidth}{
		\fontsize{12pt}{20pt}\selectfont % 
		\vspace*{0.5cm} % 

   	\hfill 	Peter von Rohr \\ Institute of Agricultural Sciences\\ D-USYS\\ ETH Zurich

		\vspace*{0.5cm} 
	}
}

\vspace*{2cm}

\fcolorbox{white}{white}{
	\parbox[t]{1.0\linewidth}{
		\centering \fontsize{25pt}{40pt}\selectfont %
		\vspace*{0.2cm}
		 751-6212-00L V \\
     Solutions to Exam    \\
     Applied Genetic Evaluation\\
     SS 2019

		\vspace*{0.7cm} % Space between the end of the title and the bottom of the grey box
	}
}


\vspace*{1cm}

<!-- % Table with Name -->
\begin{tabular}{p{3cm}p{6cm}}
Date:     &  27th May 2019 \\
          &  \\
          &  \\          
Name:     &  \\
          &  \\
          &  \\
Legi-Nr:  & \\
\end{tabular}

<!-- % Table with Points -->

\vspace{5ex}
\begin{center}
\begin{tabular}{|p{3cm}|c|c|}
\hline
Problem  &  Maximum Number of Points  &  Number of Points Reached\\
\hline
1        &  `r nPointQ1Total`  & \\
\hline
2        &  `r nPointQ2Total`  & \\
\hline
3        &  `r nPointQ3Total`  & \\
\hline
4        &  `r nPointQ4Total`  & \\
\hline
5        &  `r nPointQ5Total` & \\
\hline
Total    &  `r nPointOverallTotal` & \\
\hline
\end{tabular}
\end{center}

\vspace{1cm}

\textit{Questions in German are in italics}

\clearpage
\pagebreak

## Problem 1: 
\begin{enumerate}
\item[a)] 

\textit{}
\end{enumerate}
\points{`r lPointsQ1$TaskA`}

### Solution


\clearpage
\pagebreak

\begin{enumerate}
\item[b)] 

\textit{}
\end{enumerate}

\points{`r lPointsQ1$TaskB`}

### Solution 


\clearpage
\pagebreak

\begin{enumerate}
\item[c)] 

\textit{}
\end{enumerate}

\points{`r lPointsQ1$TaskC`}

### Solution 


\clearpage
\pagebreak

\begin{enumerate}
\item[d)] 

\textit{}
\end{enumerate}

\points{`r lPointsQ1$TaskD`}



### Solution 


\clearpage
\pagebreak

## Problem 2: 

\begin{enumerate}
\item[a)] 

\textit{}
\end{enumerate}
\points{`r lPointsQ2$TaskA`}

### Solution



\clearpage
\pagebreak

\begin{enumerate}
\item[b)] 

\textit{}
\end{enumerate}
\points{`r lPointsQ2$TaskB`}

### Solution


\clearpage
\pagebreak

\begin{enumerate}
\item[c)] 

\textit{}
\end{enumerate}
\points{`r lPointsQ2$TaskC`}



### Solution 


\clearpage
\pagebreak


## Problem 3: 

\begin{enumerate}
\item[a)] 

\textit{}
\end{enumerate}
\points{`r lPointsQ3$TaskA`}

### Solution


\clearpage
\pagebreak


\begin{enumerate}
\item[b)]

\textit{}
\end{enumerate}

\points{`r lPointsQ3$TaskB`}


### Solution

\clearpage
\pagebreak

\begin{enumerate}
\item[c)] 

\textit{}
\end{enumerate}
\points{`r lPointsQ3$TaskC`}

### Solution

\clearpage
\pagebreak


## Problem 4: 

\begin{enumerate}
\item[a)]

\textit{}
\end{enumerate}
\points{`r lPointsQ4$TaskA`}

### Solution


\clearpage
\pagebreak

\begin{enumerate}
\item[b)] 

\textit{}
\end{enumerate}
\points{`r lPointsQ4$TaskB`}

### Solution


