---
title: "Working with formula in R"
output: html_notebook
---

# Disclaimer
Have a better understanding of how to work with formulas in R


# Experiments
From https://www.datacamp.com/community/tutorials/r-formula-tutorial

```{r}
# Create variables
i <- y ~ x
j <- y ~ x + x1
k <- y ~ x + x1 + x2
```

Formulas can be joined and combined into a list

```{r}
# Join all with `c()`
l <- c(i, j, k)

# Apply `as.formula` to all elements of `f`
lapply(l, as.formula)
```

Formuals can be modified using `update()`

```{r}
(fo_current <- update(y ~ x1 + x2, ~. + x3))
```

How can we eliminate one term at the time

```{r}
(te_current <- terms(fo_current))
(vec_label_current <- labels(te_current))
# for (t in la_current){
#   fo_red <- update(fo_current, ~. - t)
#   print(fo_red)
# }
(vec_formula_current <- as.character(fo_current))
for (lidx in seq_along(vec_label_current)){
  s_cur_formula_red <- paste0(vec_label_current[-lidx], collapse = " + ")
  s_cur_formula_red <- paste0(vec_formula_current[2], vec_formula_current[1], s_cur_formula_red, collapse = " ")
  cat(" * current reduced formula: ", s_cur_formula_red, "\n")
}
```


# Application to Backwards Model Selection

We are given the following dataset

```{r}
bOnline <- FALSE
if (bOnline){
  s_data_file <- "https://charlotte-ngs.github.io/GELASMSS2019/ex/w09/data_bp_w09.csv"
} else {
  s_data_file <- file.path(here::here(), "ex/w09/data_bp_w09.csv")
}
(tbl_modsel <- readr::read_csv2(s_data_file))
```

Convert the fixed effects to factors

```{r}
tbl_modsel$sex <- as.factor(tbl_modsel$sex)
tbl_modsel$slh <- as.factor(tbl_modsel$slh)
tbl_modsel$hrd <- as.factor(tbl_modsel$hrd)
```

## Step 1: Start With Full Model
We use the full model with all predictors as the current model

```{r}
fo_current <- cw ~ sex + slh + hrd + age + day + hum
lm_current <- lm(fo_current, data = tbl_modsel)
vec_res_current <- residuals(lm_current)
n_sigmahat2_current <- crossprod(vec_res_current) / lm_current$df.residual
n_nrobs_current <- length(vec_res_current)
summary(lm_current)
```

## Step 2: Exclusion Of One Predictor
Exclude predictor that increases the residual sum of squares the least. To find the predictor to be excluded, we exclude each one of them and retain the residual sum of squares. Predictor exclusion is done via the vector of labels of the formula terms. The formulas with one predictor excluded are pasted together as strings and then converted back to formulas. We start by doing this one step at the time and then wrap it with a loop. 

```{r}
# extract the labels from the terms of the formula, resulting in a character-vector of fixed effect
(vec_formula_label <- labels(terms(fo_current)))
(vec_char_formula_current <- as.character(fo_current))
lidx <- 1
(s_formula_red <- paste0(vec_char_formula_current[2], vec_char_formula_current[1], paste0(vec_formula_label[-lidx], collapse = " + "), collapse = " "))
lm_red <- lm(as.formula(s_formula_red), data = tbl_modsel)
summary(lm_red)
n_sse_red <- crossprod(residuals(lm_red))
```

### Converting Forumulas to Strings
Interestingly enough, the `as.character()` method of a formula returns a character vector of length 3. 

```{r}
(vec_formula_char <- as.character(fo_current))
```

We need a little helper function that returns the formula as one string as

```{r}
paste0(vec_formula_char[c(2,1,3)], collapse = " ")
```

Our helper function accepts different inputs. 

1. a formula object 
2. a string specifying the response and a vector with predictors

```{r}
convert_formula_to_string <- function(pfo_model = NULL, ps_response = NULL, pvec_predictor = NULL){
  # if a formula is specified in pfo_model, then paste together the components
  if (!is.null(pfo_model)){
    vec_formula_char <- as.character(pfo_model)
    return(paste0(vec_formula_char[c(2,1,3)], collapse = " "))
  }
  # if response and predictors are given separately, paste those together
  #  start with pvec_predictor
  n_nr_predictor <- length(pvec_predictor) 
  if (n_nr_predictor > 1){
    s_pred <- paste0(pvec_predictor, collapse = " + ")
  } else if (n_nr_predictor == 1) {
      s_pred <- pvec_predictor[1]
  } else {
      s_pred <- "1"
  }
  # add response and return result
  return(paste0(ps_response, " ~ ", s_pred))
}
```

Testing the above function

```{r}
convert_formula_to_string(pfo_model = fo_current)
```
```{r}
vec_fo_current <- as.character(fo_current)
convert_formula_to_string(ps_response = vec_fo_current[2], pvec_predictor = vec_fo_current[3])
```

```{r}
vec_label <- labels(terms(fo_current))
convert_formula_to_string(ps_response = vec_fo_current[2], pvec_predictor = vec_label)
```

### Extension To All Predictors
What was done for one predictor so far is repeated for all predictors. This is done in a loop over all predictors given as strings in `vec_formual_label`.

```{r}
tbl_backstep_result <- NULL
for (lidx in seq_along(vec_formula_label)){
  s_formula_red <- paste0(vec_char_formula_current[2], vec_char_formula_current[1], paste0(vec_formula_label[-lidx], collapse = " + "), collapse = " ")
  lm_red <- lm(as.formula(s_formula_red), data = tbl_modsel)
  n_sse_red <- crossprod(residuals(lm_red))
  n_cardm_red <- length(coefficients(lm_red))
  n_cp_red <- n_sse_red / n_sigmahat2_current - n_nrobs_current  + 2 * n_cardm_red
  cat(" * lidx: ", lidx, " formula: ", s_formula_red, " sse: ", n_sse_red, " cp: ", n_cp_red, "\n")
  if (is.null(tbl_backstep_result) || n_sse_red < tbl_backstep_result$n_sse_min){
    tbl_backstep_result <- tibble::tibble( s_formula_min = as.vector(s_formula_red), 
                                           n_sse_min     = as.vector(n_sse_red),
                                           n_cp_min      = as.vector(n_cp_red) )
  }
}
cat("Predictor with min sse:\n")
print(tbl_backstep_result)
```

The result is a reduced model that has one predictor less than the full model. The predictor that was eliminated from the full model is the one that increases the residual sum of squares the least. 


### Create A Function
The above finding of the reduced model is encapsulated into a function which finds the reduced model increasing SSE the least. 

```{r}
#' Given A Model Formula, Find Model Reduced By One Predictor
#' 
#' From the given model formula pfo_model, the predictor that 
#' increases the residual sum of squares the least is eliminated 
#' resulting in the reduced model.
#'
#' @param pfo_model current model formula
#' @param ptbl_data tibble containing data
get_reduced_model <- function(pfo_model, ptbl_data){
  # convert pfo_model formula into a character vector of three elements
  # (1) "~", (2) response, (3) righthand-side of formula
  vec_char_formula_current <- as.character(pfo_model)
  # find vector of labels corresponding to predictors in pfo_model
  vec_formula_label <- labels(terms(pfo_model))
  n_nr_term <- length(vec_formula_label)
  # model that does not have predictors cannot be reduced anymore
  if (n_nr_term == 0){
    stop("[ERROR -- get_reduced_model] Cannot reduce model, because number of terms is: ", n_nr_term, "\n")
  }
  # Initialise tibble for reduced model
  tbl_backstep_result <- NULL
  # for a model with just one predictor, we construct an intercept model
  if (n_nr_term == 1){
    # s_formula_red <- paste0(vec_char_formula_current[2], 
    #                         vec_char_formula_current[1], 
    #                         "1", collapse = " ")
    s_formula_red <- convert_formula_to_string(ps_response = vec_char_formula_current[2], 
                                               pvec_predictor = c("1"))
    lm_red <- lm(as.formula(s_formula_red), data = ptbl_data)
    vec_res <- residuals(lm_red)
    n_sse_red <- crossprod(vec_res)
    tbl_backstep_result <- tibble::tibble( s_formula_min = as.vector(s_formula_red), 
                                             n_sse_min     = as.vector(n_sse_red) )
  } else {
    # loop over all predictors and find the one that increases sse the least
    for (lidx in seq_along(vec_formula_label)){
      # s_formula_red <- paste0(vec_char_formula_current[2], 
      #                         vec_char_formula_current[1], 
      #                         paste0(vec_formula_label[-lidx], collapse = " + "), 
      #                         collapse = " ")
      s_formula_red <- convert_formula_to_string(ps_response = vec_char_formula_current[2],
                                                 pvec_predictor = vec_formula_label[-lidx])
      lm_red <- lm(as.formula(s_formula_red), data = ptbl_data)
      vec_res <- residuals(lm_red)
      n_sse_red <- crossprod(vec_res)
      if (is.null(tbl_backstep_result) || n_sse_red < tbl_backstep_result$n_sse_min){
        tbl_backstep_result <- tibble::tibble( s_formula_min = as.vector(s_formula_red), 
                                               n_sse_min     = as.vector(n_sse_red) )
      }
    }
  }
  # return result
  return(tbl_backstep_result)
}
```

Testing the above function leads to

```{r}
get_reduced_model(pfo_model = fo_current, ptbl_data = tbl_modsel)
get_reduced_model(pfo_model = cw~sex, ptbl_data = tbl_modsel)
# get_reduced_model(pfo_model = cw~1, ptbl_data = tbl_modsel)

```

The second function is used to compute the $C_p$ value for a given model. 

```{r}
#' Compute Mallows C_p Statistic for model pfo_model
compute_cp <- function(pfo_model, pn_sigmahat2_full_model, ptbl_data){
  # fit the model given by pfo_model to data in ptbl_data
  lm_model <- lm(pfo_model, data = ptbl_data)
  # get residuals
  vec_res <- residuals(lm_model)
  # get number of observations
  n_nr_obs <- length(vec_res)
  # get residual sum of squares
  n_sse <- crossprod(vec_res)
  # get number of fixed effects
  n_cardm <- length(coefficients(lm_model))
  # compute cp
  n_cp_result <- n_sse / pn_sigmahat2_full_model - n_nr_obs + 2 * n_cardm
  # return result
  return(n_cp_result)
}
```

The above function is tested with

```{r}
compute_cp(pfo_model = fo_full, pn_sigmahat2_full_model = n_sigmahat2_current, ptbl_data = tbl_modsel)
```

Reducing the full model by one predictor is done with `get_reduced_model()`

```{r}
(tbl_model_red <- get_reduced_model(pfo_model = fo_full, ptbl_data = tbl_modsel))
```

The $C_p$ value for the reduced model is then

```{r}
compute_cp(pfo_model = as.formula(tbl_model_red$s_formula_min), pn_sigmahat2_full_model = n_sigmahat2_current, ptbl_data = tbl_modsel)
```



## Step 3: Repeat Step 2 Until All Predictors Are Eliminated
Starting from the full model, we repeat step 2 until we have the model with only an intercept. This produces a sequence of models. From this sequence, we select the one with the smallest $C_p$ value. 

We construct a function that takes the full model as a formula and from that generates the sequence of models. 

```{r}
# unroll the function below
(pfo_full_model <- fo_full)
ptbl_data <- tbl_modsel
```

```{r}
construct_model_sequences <- function(pfo_full_model, ptbl_data){
  # initialise a tibble that will hold the sequence of models
  tbl_model_sequence_result <- NULL
  # start with a fit of the full model and compute estimate of error variance
  fo_current_model <- pfo_full_model
  lm_current_model <- lm(fo_current_model, data = ptbl_data)
  n_sigmahat2_full_model <- crossprod(residuals(lm_current_model)) / lm_current_model$df.residual
  # extract lables of model terms for full model corresponding 
  #  to all predictors in a character vector
  vec_label_current_model <- labels(terms(lm_current_model))
  n_nr_pred_current_model <- length(vec_label_current_model)
  # loop until there are any predictors in the model
  while (n_nr_pred_current_model > 0){
    # compute C_p for current model and add it to result tibble
    n_cp_current_model <- compute_cp(pfo_model               = fo_current_model, 
                                     pn_sigmahat2_full_model = n_sigmahat2_full_model,
                                     ptbl_data               = ptbl_data)
    
    # convert model into a character vector
    vec_fo_current_model <- as.character(fo_current_model)
    # put information about current model into a tibble
    tbl_cur_model <- tibble::tibble(model = as.vector(convert_formula_to_string(ps_response    = vec_fo_current_model[2],
                                                                      pvec_predictor = vec_label_current_model)),
                                    cp    = as.vector(n_cp_current_model))
    # add current model to result structure
    if (is.null(tbl_model_sequence_result)){
      tbl_model_sequence_result <- tbl_cur_model
    } else {
      tbl_model_sequence_result <- dplyr::bind_rows(tbl_model_sequence_result, tbl_cur_model)
    }
    # reduce current model by one predictor
    tbl_red_model <- get_reduced_model(pfo_model    = fo_current_model,
                                       ptbl_data    = ptbl_data)
    # update
    fo_current_model <- as.formula(tbl_red_model$s_formula_min)
    lm_current_model <- lm(fo_current_model, data = ptbl_data)
    vec_label_current_model <- labels(terms(lm_current_model))
    n_nr_pred_current_model <- length(vec_label_current_model)
  }
  # add the intercept model
  # compute C_p for current model and add it to result tibble
  n_cp_current_model <- compute_cp(pfo_model               = fo_current_model, 
                                   pn_sigmahat2_full_model = n_sigmahat2_full_model,
                                   ptbl_data               = ptbl_data)
  
  # convert model into a character vector
  vec_fo_current_model <- as.character(fo_current_model)
  # put information about current model into a tibble
  tbl_cur_model <- tibble::tibble(model = as.vector(convert_formula_to_string(ps_response    = vec_fo_current_model[2],
                                                                              pvec_predictor = vec_label_current_model)),
                                  cp    = as.vector(n_cp_current_model))
  # add current model to result structure
  if (is.null(tbl_model_sequence_result)){
    tbl_model_sequence_result <- tbl_cur_model
  } else {
    tbl_model_sequence_result <- dplyr::bind_rows(tbl_model_sequence_result, tbl_cur_model)
  }

  return(tbl_model_sequence_result)  
}

construct_model_sequences(pfo_full_model = fo_full, ptbl_data = tbl_modsel)
```



