---
title: "Data Generation For Breeding Program Analysis"
date: "2019-04-14"
author: "Peter von Rohr"
output: html_notebook
---

```{r}
set.seed(6576)
```

# Disclaimer
Exercise 1 in Applied Genetic Evaluation (GEL) does a genetic evaluation of a beef cattle data set. 


# Important Factors
According to the work by Kunz et al (2019) presented at SVT (16.04.2019) important factors for the carcass performance traits are

* sex
* slaughterhouse
* classifier
* age (as linear and quadratic covariable)
* herd * year (as random interaction)


# Factors Used In Simulation
For our data simulation, we use a simplified set of factors consisting of 

* sex
* slaughterhouse
* age as linear covariable
* herd as fixed effect.


# Trait
For reasons of simplicity we are using carcass weight, because it is a continuous trait.


# Pedigree
We use `XSim` to generate a pedigree. The output pedigree is stored under `data_w09.ped`. It is read into a `tibble` with 

```{r}
s_ped_file <- "data_w09.ped"
tbl_ped <- readr::read_delim(file = s_ped_file, delim = " ", col_names = FALSE)
### keep only first three columns
tbl_ped <- tbl_ped[,1:3]
tbl_ped
```


# Data Components
## Sex
Add the sex to the pedigree and use it for data

```{r}
require(dplyr)
tbl_data <- tibble(Id = as.numeric(tbl_ped[[1]]))
tbl_data <- bind_cols(tbl_data, tibble(sex = rep(2, nrow(tbl_data))))
# set the sex of all fathers to 1
# find ids of fathers
vec_father_ids <- setdiff(as.numeric(unique(tbl_ped$X2)), 0)
tbl_data$sex[vec_father_ids] <- 1
table(tbl_data$sex)
```

## Slaughterhouses
We randomly assign three slaughterhouses

```{r}
n_nr_slh <- 3
vec_slh <- floor(n_nr_slh * runif(nrow(tbl_data))) + 1
table(vec_slh)
tbl_data <- bind_cols(tbl_data, tibble(slh = vec_slh))
```

## Herds
Randomly assign five herds

```{r}
n_nr_hrd <- 5
vec_hrd <- floor(n_nr_hrd * runif(nrow(tbl_data))) + 1
table(vec_hrd)
tbl_data <- bind_cols(tbl_data, tibble(hrd = vec_hrd))
```

## Age
Normal distribution for the age

```{r}
n_age_mean <- 400
n_age_sd <- 5.7
vec_age <- rnorm(n = nrow(tbl_data), mean = n_age_mean, sd = n_age_sd)
tbl_data <- bind_cols(tbl_data, tibble(age = round(vec_age, digits = 0)))
```


# Effects

## sex
The effect of sex is encoded in a vector of length two

```{r}
vec_sex_eff <- c(50, -24)
```

The effects of all the animals are just the contribution added

```{r}
# design matrix
mat_sex <- matrix(0, nrow = nrow(tbl_data), ncol = 2)
# males
mat_sex[which(tbl_data$sex == 1),1] <- 1
# females
mat_sex[which(tbl_data$sex == 2),2] <- 1
mat_sex_eff <- crossprod(t(mat_sex), vec_sex_eff)
```

## slaughterhouse
```{r}
vec_slh_eff <- c(-7, 15, -3)
mat_slh <- matrix(0, nrow = nrow(tbl_data), ncol = length(vec_slh_eff))
for (slh_idx in seq_along(vec_slh_eff)){
  mat_slh[which(tbl_data$slh == slh_idx), slh_idx] <- 1
}
mat_slh_eff <- crossprod(t(mat_slh), vec_slh_eff)
```


## herd

```{r}
vec_hrd_eff <- c(-23, 65, -14, 36, -3)
mat_hrd <- matrix(0, nrow = nrow(tbl_data), ncol = length(vec_hrd_eff))
for (hrd_idx in seq_along(vec_hrd_eff)){
  mat_hrd[which(tbl_data$hrd == hrd_idx), hrd_idx] <- 1
}
mat_hrd_eff <- crossprod(t(mat_hrd), vec_hrd_eff)
```


## age as linear covariable
```{r}
n_age_reg <- 0.632
vec_age_eff <- (tbl_data$age - mean(tbl_data$age)) * n_age_reg
```


## Random error
Errors are sampled as random normal deviate with a fixed standard deviation

```{r}
n_sd_error <- 7.43
vec_rand_error <- rnorm(n = nrow(tbl_data), mean = 0, sd = n_sd_error)
```

# Compose the data set
Summing up all the effects leads to

```{r}
n_inter <- 250
mat_y <- n_inter + mat_sex_eff + mat_slh_eff + mat_hrd_eff + vec_age_eff + vec_rand_error
```

Statistics about observations

```{r}
mean(mat_y)
```

```{r}
sd(mat_y)
```

Adding the observations to the data 

```{r}
tbl_data <- bind_cols(tbl_data, tibble(cw=round(mat_y[,1], digits = 0)))
tbl_data
```

# Adding two random components

```{r}
vec_day <- floor(30*runif(nrow(tbl_data))) + 1
vec_hum <- rnorm(nrow(tbl_data))
tbl_data <- bind_cols(tbl_data, tibble(day=vec_day, hum = round(vec_hum, digits = 1)))
                      
```

# Output to csv

```{r}
s_data_file <- "data_bp_w09.csv"
readr::write_csv2(tbl_data, path = s_data_file)
```

