---
title: "Data Generation For Breeding Program Analysis"
date: "2019-04-14"
author: "Peter von Rohr"
output: html_notebook
---

```{r}
set.seed(6576)
```

# Disclaimer
Exercise 1 in Applied Genetic Evaluation (GEL) does a genetic evaluation of a beef cattle data set. 


# Important Factors
According to the work by Kunz et al (2019) presented at SVT (16.04.2019) important factors for the carcass performance traits are

* sex
* slaughterhouse
* classifier
* age (as linear and quadratic covariable)
* herd * year (as random interaction)


# Factors Used In Simulation
For our data simulation, we use a simplified set of factors consisting of 

* sex
* slaughterhouse
* age as linear covariable
* herd as fixed effect.


# Trait
For reasons of simplicity we are using carcass weight, because it is a continuous trait.


# Pedigree
We use `XSim` to generate a pedigree. The output pedigree is stored under `data_w09.ped`. It is read into a `tibble` with 

```{r}
s_ped_file <- "data_w09.ped"
tbl_ped <- readr::read_delim(file = s_ped_file, delim = " ", col_names = FALSE)
### keep only first three columns and order according to id
tbl_ped <- tbl_ped[,1:3]
names(tbl_ped) <- c("id", "sire", "dam")
tbl_ped$id <- as.integer(tbl_ped$id)
tbl_ped <- tbl_ped[order(tbl_ped$id),]
tbl_ped
```


# Data Components
## Sex
Add the sex to the pedigree and use it for data

```{r}
require(dplyr)
tbl_data <- tibble(Id = as.integer(tbl_ped[[1]]))
tbl_data <- bind_cols(tbl_data, tibble(sex = rep(2, nrow(tbl_data))))
# set the sex of all fathers to 1
# find ids of fathers
vec_father_ids <- setdiff(as.numeric(unique(tbl_ped$X2)), 0)
tbl_data$sex[vec_father_ids] <- 1
table(tbl_data$sex)
```

## Slaughterhouses
We randomly assign three slaughterhouses

```{r}
n_nr_slh <- 3
vec_slh <- floor(n_nr_slh * runif(nrow(tbl_data))) + 1
table(vec_slh)
tbl_data <- bind_cols(tbl_data, tibble(slh = vec_slh))
```

## Herds
Randomly assign five herds

```{r}
n_nr_hrd <- 5
vec_hrd <- floor(n_nr_hrd * runif(nrow(tbl_data))) + 1
table(vec_hrd)
tbl_data <- bind_cols(tbl_data, tibble(hrd = vec_hrd))
```

## Age
Normal distribution for the age

```{r}
n_age_mean <- 400
n_age_sd <- 5.7
vec_age <- rnorm(n = nrow(tbl_data), mean = n_age_mean, sd = n_age_sd)
tbl_data <- bind_cols(tbl_data, tibble(age = round(vec_age, digits = 0)))
```


# Effects

## sex
The effect of sex is encoded in a vector of length two

```{r}
vec_sex_eff <- c(50, -24)
```

The effects of all the animals are just the contribution added

```{r}
# design matrix
mat_sex <- matrix(0, nrow = nrow(tbl_data), ncol = 2)
# males
mat_sex[which(tbl_data$sex == 1),1] <- 1
# females
mat_sex[which(tbl_data$sex == 2),2] <- 1
mat_sex_eff <- crossprod(t(mat_sex), vec_sex_eff)
```

## slaughterhouse
```{r}
vec_slh_eff <- c(-7, 15, -3)
mat_slh <- matrix(0, nrow = nrow(tbl_data), ncol = length(vec_slh_eff))
for (slh_idx in seq_along(vec_slh_eff)){
  mat_slh[which(tbl_data$slh == slh_idx), slh_idx] <- 1
}
mat_slh_eff <- crossprod(t(mat_slh), vec_slh_eff)
```


## herd

```{r}
vec_hrd_eff <- c(-23, 65, -14, 36, -3)
mat_hrd <- matrix(0, nrow = nrow(tbl_data), ncol = length(vec_hrd_eff))
for (hrd_idx in seq_along(vec_hrd_eff)){
  mat_hrd[which(tbl_data$hrd == hrd_idx), hrd_idx] <- 1
}
mat_hrd_eff <- crossprod(t(mat_hrd), vec_hrd_eff)
```


## age as linear covariable
```{r}
n_age_reg <- 0.632
vec_age_eff <- (tbl_data$age - mean(tbl_data$age)) * n_age_reg
```


## Random error
Errors are sampled as random normal deviate with a fixed standard deviation

```{r}
n_sd_error <- 7.43
vec_rand_error <- rnorm(n = nrow(tbl_data), mean = 0, sd = n_sd_error)
```


## Polygenic Component
In addition to the fixed effects and in addition to the random error, we add a polygenic component. 

```{r}
# polygenic standard deviation
n_sd_gen <- 6.8
# change missing value code in tbl_ped from 0 to NA
tbl_ped$sire <- as.integer(tbl_ped$sire)
tbl_ped$sire[tbl_ped$sire == 0L] <- NA
tbl_ped$dam <- as.integer(tbl_ped$dam)
tbl_ped$dam[tbl_ped$dam == 0L] <- NA
tbl_ped$id <- as.integer(tbl_ped$id)
# plausiblity check for pedigree
#sum((tbl_ped$id - order(tbl_ped$id))^2)
head(which(tbl_ped$id != order(tbl_ped$id)))
head(tbl_ped, 20)
```

```{r}
which(tbl_ped$id == 49)
```

```{r}
tbl_ped[74, ]
```



```{r}
# construct the numerator relationship matrix using pedigreemm
ped <- pedigreemm::pedigree(sire  = tbl_ped$sire, 
                            dam   = tbl_ped$dam,
                            label = tbl_ped$id)
matA <- pedigreemm::getA(ped)
```

Do a cholesky-decomposition of `matA` to be able to simulate the breeding values.

```{r}
matR <- chol(matA)
```

Now generate a vector of random number

```{r}
vec_rand_polygen <- rnorm(n = nrow(tbl_ped), mean = 0, sd = n_sd_gen)
vec_bv <- crossprod(matR, vec_rand_polygen)
mean(vec_bv);sd(vec_bv)
```


# Compose the data set
Summing up all the effects leads to

```{r}
n_inter <- 250 - mean(vec_bv)
mat_y <- n_inter + mat_sex_eff + mat_slh_eff + mat_hrd_eff + vec_age_eff + vec_rand_error + vec_bv
```

Statistics about observations

```{r}
mean(mat_y)
```

```{r}
sd(mat_y)
```

Adding the observations to the data 

```{r}
tbl_data <- bind_cols(tbl_data, tibble(cw=round(mat_y[,1], digits = 0)))
tbl_data_relevant_eff <- bind_cols(tbl_data, 
                                   tibble(sire=tbl_ped$sire),
                                   tibble(dam=tbl_ped$dam))
```

# Adding two random components

```{r}
vec_day <- floor(30*runif(nrow(tbl_data))) + 1
vec_hum <- rnorm(nrow(tbl_data))
tbl_data <- bind_cols(tbl_data, tibble(day=vec_day, hum = round(vec_hum, digits = 1)))
                      
```

# Output to csv

```{r}
s_data_file <- "data_bp_w09.csv"
readr::write_csv2(tbl_data, path = s_data_file)
# relevant effect
s_data_relevant_eff <- "data_bp_w12.csv"
readr::write_csv2(tbl_data_relevant_eff, path = s_data_relevant_eff)
```


# Extract Sire Model Data

```{r}
# (tbl_sire <- tbl_data_relevant_eff[tbl_data_relevant_eff$sire <= 10 & ,])
library(dplyr)
tbl_data_relevant_eff %>%
  filter(!is.na(sire)) %>%
  filter(sire < 11) %>%
  select(Id, slh, hrd, age, cw, sire) -> tbl_sire
```

Write the data to a file

```{r}
s_data_file_sire <- "data_sire_w12.csv"
readr::write_csv2(tbl_sire, path = s_data_file_sire)
```


We try to fit an ANOVA using the sire model data

```{r}
tbl_sire$slh <- as.factor(tbl_sire$slh)
tbl_sire$hrd <- as.factor(tbl_sire$hrd)
tbl_sire$sire <- as.factor(tbl_sire$sire)
aov_sire <- aov(cw ~ slh + hrd + age + sire, data = tbl_sire)
(sum_aov_sire <- summary(aov_sire))
```


The counts per sire

```{r}
tbl_sire %>%
  group_by(sire) %>%
  summarise(prog_count = n())
```

