---
title: Applied Genetic Evaluation - Solution 2
author: Peter von Rohr
date: "2019-05-06"
output:
  bookdown::pdf_document2:
    toc: false
    number_sections: false
    keep_tex: false
header-includes:
 \usepackage{longtable}
 \usepackage{float}
---

```{r setup, include=FALSE}
# knitr::opts_chunk$set(echo = FALSE, results = 'asis', fig.pos = 'H')
knitr::knit_hooks$set(hook_convert_odg = rmddochelper::hook_convert_odg)
```

## Problem 1: Variance Components Estimation
We are given the dataset with the response variable `carcass weight` (CW) and the predictor variables that resulted from the model selection process from Exercise 1. These consisted of

* sex (`sex`)
* slaughterhouse (`slh`)
* herd (`hrd`)
* age at slaughter (`age`)

The data is available from `https://charlotte-ngs.github.io/GELASMSS2019/ex/w12/data_bp_w12.csv`. 

We use a mixed linear effects model to estimate the variance components for the random effects in the model.

\begin{equation}
y = Xb + Za + e
\end{equation}

where $y$ is a vector of observations, $b$ is a vector of fixed effects found to be relevant in Exercise 1, $a$ is a vector of random breeding values and $e$ is a vector of random errors. 


### Hint
* Use the package `pedigreemm` to get an estimate of the variance components


### Your Task
* Estimate the variance components $\sigma_a^2$ and $\sigma_e^2$ for the two random component $a$ and $e$, respectively.


### Solution
We first have to read the data

```{r}
s_data_path_gel_ex2 <- "https://charlotte-ngs.github.io/GELASMSS2019/ex/w12/data_bp_w12.csv"
tbl_gel_ex2 <- readr::read_csv2(file = s_data_path_gel_ex2)
colnames(tbl_gel_ex2);dim(tbl_gel_ex2)
```

The fixed effects are converted into factors

```{r}
tbl_gel_ex2$sex <- as.factor(tbl_gel_ex2$sex)
tbl_gel_ex2$slh <- as.factor(tbl_gel_ex2$slh)
tbl_gel_ex2$hrd <- as.factor(tbl_gel_ex2$hrd)
```

From the help file of `pedigreemm`, we can see that we first have to define a pedigree. 

```{r}
ped <- pedigreemm::pedigree(sire  = tbl_gel_ex2$sire, 
                            dam   = tbl_gel_ex2$dam,
                            label = tbl_gel_ex2$Id)
```

Now the model can be specified as for the other functions to fit linear mixed effects model, such as `lmer`. 

```{r}
require(pedigreemm)
# according to https://stat.ethz.ch/pipermail/r-sig-mixed-models/2014q1/021609.html
options(lmerControl=list(check.nobs.vs.nlev="ignore",
     check.nobs.vs.rankZ = "ignore",
     check.nobs.vs.nRE="ignore"))
lmem_gel_ex2 <- pedigreemm(cw ~ sex + slh + hrd + age + (1|Id), data = tbl_gel_ex2, pedigree = list(Id = ped))
# lmem_gel_ex2 <- pedigreemm(cw ~ (1|Id), data = tbl_gel_ex2, pedigree = list(Id = ped))
summary(lmem_gel_ex2)
```









