---
title: Applied Genetic Evaluation - Solution 2
author: Peter von Rohr
date: "2019-05-06"
output:
  bookdown::pdf_document2:
    toc: false
    number_sections: false
    keep_tex: false
header-includes:
 \usepackage{longtable}
 \usepackage{float}
---

```{r setup, include=FALSE}
# knitr::opts_chunk$set(echo = FALSE, results = 'asis', fig.pos = 'H')
knitr::knit_hooks$set(hook_convert_odg = rmddochelper::hook_convert_odg)
```

## Problem 1: Analysis of Variance
Estimate the variance component for the sire effect using an analysis of variance. The data is available from `https://charlotte-ngs.github.io/GELASMSS2019/ex/w12/data_sire_w12.csv`. Because the data contains just female animals, the fixed effect of the sex can no longer be estimated. 


### Hint
* Use the functions `aov()` to do the analysis of variance and the function `summary()` on the ANOVA result to get the relevant parts of the variance components.


### Solution
The data is read using 

```{r}
s_data_sire <- "https://charlotte-ngs.github.io/GELASMSS2019/ex/w12/data_sire_w12.csv"
tbl_sire_aov <- readr::read_csv2(file = s_data_sire)
tbl_sire_aov$slh <- as.factor(tbl_sire_aov$slh)
tbl_sire_aov$hrd <- as.factor(tbl_sire_aov$hrd)
tbl_sire_aov$sire <- as.factor(tbl_sire_aov$sire)
# anova
aov_sire <- aov(cw ~ slh + hrd + age + sire, data = tbl_sire_aov)
(summary_aov_sire <- summary(aov_sire))
```

From the Mean Sq of the residuals and of the sires, the estimates of the variance components can be computed. 

The estimate $\hat{\sigma}_e^2$ corresponds to the mean sum of squares for the residuals. Hence 

```{r, echo=FALSE}
n_hatsigmae2 <- summary_aov_sire[[1]]$`Mean Sq`[5]
```

$$\hat{\sigma}_e^2 = `r round(n_hatsigmae2, 0)`$$

The  estimate of the sire variance $\hat{\sigma}_s^2$ is computed as 

$$\hat{\sigma}_s^2 = \frac{Mean\ Sq(sire) -  \hat{\sigma}_e^2}{k}$$

where $k = \frac{1}{r-1} \left[ N - \frac{\sum_{i=1}^{r}n_i^2}{N} \right]$ with $r$ is the number of sires, $N$ the total number of observations and $n_i$ the number of progeny for sire $i$. To compute the value $k$, we need the progeny counts for each sire. 

```{r, echo=FALSE, message=FALSE}
library(dplyr)
tbl_sire_aov %>%
    group_by(sire) %>%
    summarise(prog_count = n()) -> tbl_prog_count
n_nr_sire <- nrow(tbl_prog_count)
n_nr_obs <- nrow(tbl_sire_aov)
n_sum_ni2 <- sum(tbl_prog_count$prog_count^2)
n_k <- 1/(n_nr_sire-1) * (n_nr_obs - n_sum_ni2 / n_nr_obs)
knitr::kable(tbl_prog_count, booktabs = TRUE, longtable = TRUE)
```

From the above table, we get

$$k = \frac{1}{r-1} \left[ N - \frac{\sum_{i=1}^{r}n_i^2}{N} \right] = 1/(`r n_nr_sire`- 1) \left[`r n_nr_obs` - \frac{`r n_sum_ni2`}{`r n_nr_obs`}  \right] = `r round(n_k, 1)`$$

```{r}

n_msqsire <- summary_aov_sire[[1]]$`Mean Sq`[4]
n_hatsigmas2 <- (n_msqsire - n_hatsigmae2) / n_k
```

Putting everything together, we get

$$\hat{\sigma}_s^2 = \frac{`r round(n_msqsire, 0)` -  `r round(n_hatsigmae2, 0)`}{`r round(n_k, 1)`} = `r round(n_hatsigmas2, 1)`$$

From the estimate of the sire variance, we get the estimate of the genetic variance by multiplying it with four. Hence

$$\hat{\sigma}_a^2 = 4 * \hat{\sigma}_s^2 = 4 * `r round(n_hatsigmas2, 1)` = `r round(4*n_hatsigmas2, 1)`$$


## Problem 2: Variance Components Estimation Using REML
Use the same data set as for Problem 1 and a sire model to estimate the same sire variance $\sigma_s^2$. The sire model is the linear mixed effects model that contains sire effects as random component. The model can be specified as 

$$y = Xb + Zs + e$$

where $y$ is the vector of observations, $b$ is the vector of fixed effects which are the same as in Problem 1, $s$ is the vector of random sire effects and $e$ is the vector of random error terms. 

### Hint 
* Use the package `pedigreemm` to get a REML estimate for the sire variance component $\sigma_s^2$. 
* We assume that the sires are not related. Hence variance-covariance matrix $var(s)$ of the sire components are $var(s) = I * \sigma_s^2$. 


### Solution
As the first step, we have to specify a pedigree. The sires are unrelated hence the corresponding pedigree corresponds to

```{r}
ped_sire <- pedigreemm::pedigree(sire = rep(NA, n_nr_sire), 
                                 dam = rep(NA, n_nr_sire), 
                                 label = c(1:n_nr_sire))
```

The specified pedigree is used for the linear mixed effects model

```{r}
require(pedigreemm)
lmem_sire <- pedigreemm(cw ~ slh + hrd + age + (1|sire), 
                        pedigree = list(sire = ped_sire), 
                        data = tbl_sire_aov)
summary(lmem_sire)
```




## Additional Problem: Variance Components Estimation Using an Animal Model
We are given the dataset with the response variable `carcass weight` (CW) and the predictor variables that resulted from the model selection process from Exercise 1. These consisted of

* sex (`sex`)
* slaughterhouse (`slh`)
* herd (`hrd`)
* age at slaughter (`age`)

The data is available from `https://charlotte-ngs.github.io/GELASMSS2019/ex/w12/data_bp_w12.csv`. 

We use a mixed linear effects model to estimate the variance components for the random effects in the model.

\begin{equation}
y = Xb + Za + e
\end{equation}

where $y$ is a vector of observations, $b$ is a vector of fixed effects found to be relevant in Exercise 1, $a$ is a vector of random breeding values and $e$ is a vector of random errors. 


### Hint
* Use the package `pedigreemm` to get an estimate of the variance components


### Your Task
* Estimate the variance components $\sigma_a^2$ and $\sigma_e^2$ for the two random component $a$ and $e$, respectively.


### Solution
We first have to read the data

```{r}
s_data_path_gel_ex2 <- "https://charlotte-ngs.github.io/GELASMSS2019/ex/w12/data_bp_w12.csv"
tbl_gel_ex2 <- readr::read_csv2(file = s_data_path_gel_ex2)
colnames(tbl_gel_ex2);dim(tbl_gel_ex2)
```

The fixed effects are converted into factors

```{r}
tbl_gel_ex2$sex <- as.factor(tbl_gel_ex2$sex)
tbl_gel_ex2$slh <- as.factor(tbl_gel_ex2$slh)
tbl_gel_ex2$hrd <- as.factor(tbl_gel_ex2$hrd)
```

From the help file of `pedigreemm`, we can see that we first have to define a pedigree. 

```{r}
ped <- pedigreemm::pedigree(sire  = tbl_gel_ex2$sire, 
                            dam   = tbl_gel_ex2$dam,
                            label = tbl_gel_ex2$Id)
```

Now the model can be specified as for the other functions to fit linear mixed effects model, such as `lmer`. 

```{r eval=FALSE}
# This takes more than one hour to run.
require(pedigreemm)
# according to https://stat.ethz.ch/pipermail/r-sig-mixed-models/2014q1/021609.html
options(lmerControl=list(check.nobs.vs.nlev="ignore",
     check.nobs.vs.rankZ = "ignore",
     check.nobs.vs.nRE="ignore"))
s_lmem_file <- "lmem_gel_ex2.rds"
if (file.exists(s_lmem_file)){
  load(file = s_lmem_file)
} else {
  lmem_gel_ex2 <- pedigreemm(cw ~ sex + slh + hrd + age + (1|Id), 
                             data = tbl_gel_ex2, 
                             pedigree = list(Id = ped))
  saveRDS(lmem_gel_ex2, file = s_lmem_file)
}
summary(lmem_gel_ex2)
```









