---
title: "Compute Mallow $C_p$"
output: html_notebook
---

# Disclaimer
Show the how to compute the Mallow $C_p$ statistic using the BW on BC regression data.


# Example Data

```{r}
set.seed(432)
suppressPackageStartupMessages( require(dplyr) )
tbl_reg <- tibble::tibble(`Breast Circumference` = c(176, 177, 178, 179, 179, 180, 181, 182,183, 184),
                          `Body Weight` = c(471, 463, 481, 470, 496, 491, 518, 511, 510, 541))
n_nr_ani <- nrow(tbl_reg)
tbl_reg <- bind_cols(tibble::tibble(Animal = 1:n_nr_ani),
                     tbl_reg)
vec_randpred <- runif(n_nr_ani, 
                      min = min(tbl_reg$`Breast Circumference`), 
                      max = max(tbl_reg$`Breast Circumference`))
tbl_reg_aug <- bind_cols(tbl_reg, tibble(RandPred = round(vec_randpred, digits = 0)))
colnames(tbl_reg_aug) <- c("Animal", "BC", "BW", "RandPred")
tbl_reg_aug
```


# Computation of $C_p$
The Mallow $C_p$ statistic is computed as 

$$C_p(\mathcal{M}) = \frac{SSE(\mathcal{M})}{\hat{\sigma}^2} - n + 2 |\mathcal{M}|$$

where $\hat{\sigma}^2$ is the estimate of the error variance under the full model. Hence we need the fit of the full model first.

```{r}
lm_full <- lm(BW ~ BC + RandPred, data = tbl_reg_aug)
vec_res_full <- residuals(lm_full)
n_sse_full <- crossprod(vec_res_full)
sqrt(n_sigma_hat2 <- n_sse_full / lm_full$df.residual)
summary(lm_full)
n_cardm_full <- length(coefficients(lm_full))
(n_cp_full <- n_sse_full / n_sigma_hat2 - n_nr_ani + 2 * n_cardm_full)
```

For the reduced model

```{r}
lm_red <- lm(BW ~ BC, data = tbl_reg_aug)
vec_res_red <- residuals(lm_red)
n_sse_red <- crossprod(vec_res_red)
n_cardm_red <- length(coefficients(lm_red))
(n_cp_red <- n_sse_red / n_sigma_hat2 - n_nr_ani + 2 * n_cardm_red)
```


# Verify with `olsrr`
The $C_p$ value can also be computed with `olsrr`

```{r}
library(olsrr)
ols_step_best_subset(lm_full)
```


