---
title: |
  |  Genetic Evaluation of Lifestock and
  |  Applied Statistical Methods in Animal Science
author: "Peter von Rohr"
date: "2018-12-31"
output: 
  html_document:
    toc_float: true
    toc: true
    theme: united
    highlight: tango
---

```{r CourseSetup, echo=FALSE, results='hide'}
# BaseUrl of course website
sBaseUrl <- 'https://charlotte-ngs.github.io/GELASMSS2019/'
# define course start date
dCourseStart <- as.Date("2019/02/18")
# set number of weeks in semester
nNrSemesterWeeks <- 15
# define columns for weaks and dates which remain constant
Woche <- 1:nNrSemesterWeeks
Datum <- format(seq(dCourseStart, by = "week", length.out = nNrSemesterWeeks), "%d.%m")
# empty vector for topics
vec_topics <- vector(mode = "character", length = nNrSemesterWeeks)
# set subject for special dates
# holiday
sEasterMon <- format(as.Date("2019/04/22"), "%d.%m")
vec_topics[which(Datum == sEasterMon)] <- "__Easter Monday__"
# exams
vec_topics[which(Datum == format(as.Date("2019/05/27"), "%d.%m"))] <- "__Final Exams__"
```

__Welcome__ to 

- [751-7602-00L	Applied Statistical Methods in Animal Sciences](http://www.vorlesungsverzeichnis.ethz.ch/Vorlesungsverzeichnis/lerneinheit.view?semkez=2019S&ansicht=ALLE&lerneinheitId=128454&lang=en)
- [751-6212-00L Genetic Evaluation of Lifestock](http://www.vorlesungsverzeichnis.ethz.ch/Vorlesungsverzeichnis/lerneinheit.view?semkez=2019S&ansicht=ALLE&lerneinheitId=127426&lang=en)


## Topics for Master Theses
Topics for Master Theses at Qualitas AG together with ETH are shown [here](`r paste0(sBaseUrl, 'misc/20181231_master_thesis_topics/MasterThesisTopics_SS2019.html')`).


## Description
The following descriptions are taken from the course catalog

### Applied Statistical Methods in Animal Science
Genomic selection is currently the method of choice for improving the genetic potential of selection candidates in livestock breeding programs. This lecture introduces the reason why regression cannot be used in genomic selection. Alternatives to regression analysis that are suitable for genomic selection are presented. The concepts introduced are illustrated by excersises in R.

The students are familiar with the properties of multiple linear regression and they are able to analyse simple data sets using regression methods. The students know why multiple linear regression cannot be used for genomic selection. The students know the statistical methods used in genomic selection, such as BLUP-based approaches, Bayesian procedures and LASSO. The students are able to solve simple exercise problems using the statistical framework R.

### Genetic Evaluation of Lifestock
Swiss routine breeding value estimation/genetic evaluation systems of cattle, pig, sheep and goats are presented with methods and evaluated traits. Examples will be demonstrated using the statistical software R.

The students know the theoretical and practical application of breeding value estimation in Switzerland for cattle, pig, sheep and goats. The students are able to interpret estimated breeding values.


## Information
- Lecturer: Peter von Rohr
- Date: Montag 8-10
- Room: [LFW C11](http://www.mapsearch.ethz.ch/map/map.do?gebaeudeMap=LFW&lang=en)

<!--
### Excursion 
More information on the excursion will be available from here. [here](`r paste0(sBaseUrl, "misc/gel_exkursion/gel_exkursion.html") `)-->


### Exercise Platform
You can use the exercise platform to work on exercise problems. The platform is reachable at: http://r4tea.rteastem.org:8787/

The username corresponds to the part of your E-Mail address before the `@`-sign. The password is sent to you via a different E-Mail.


## Final Exams
- Applied Statistical Methods: 27.05, 08:15 - 09:00 
- Genetic Evaluation of Livestock: 27.05, 09:15 - 10:00


## Questions
You can always ask questions during the lecture and the exercise hours or alternatively via e-mail at (peter.vonrohr `at` gmail.com). 


## Schedule
```{r Schedule, echo=FALSE, results='asis'}
# define vector of topics for course schedule
vec_topics_schedule <- vec_topics
vec_topics_schedule[which(Datum == format(as.Date("2019/02/18"), "%d.%m"))] <- "Introduction"
vec_topics_schedule[which(Datum == format(as.Date("2019/02/25"), "%d.%m"))] <- "Multiple Linear Regression"
vec_topics_schedule[which(Datum == format(as.Date("2019/03/04"), "%d.%m"))] <- "GBLUP - Marker-Effects Models"
vec_topics_schedule[which(Datum == format(as.Date("2019/03/11"), "%d.%m"))] <- "GBLUP - Breeding Value Models"
vec_topics_schedule[which(Datum == format(as.Date("2019/03/18"), "%d.%m"))] <- "Lasso"
vec_topics_schedule[which(Datum == format(as.Date("2019/03/25"), "%d.%m"))] <- "Bayesian Approaches"
### # ========================================================= ###
vec_topics_schedule[which(Datum == format(as.Date("2019/04/01"), "%d.%m"))] <- "Introduction"
vec_topics_schedule[which(Datum == format(as.Date("2019/04/08"), "%d.%m"))] <- "Suisag and The Swiss Pig Breeding Program - Patrick Stratz (Suisag)"
vec_topics_schedule[which(Datum == format(as.Date("2019/04/15"), "%d.%m"))] <- "Model Selection"
vec_topics_schedule[which(Datum == format(as.Date("2019/04/29"), "%d.%m"))] <- "Genetic Groups"
vec_topics_schedule[which(Datum == format(as.Date("2019/05/06"), "%d.%m"))] <- "Longitudinal Data"
vec_topics_schedule[which(Datum == format(as.Date("2019/05/13"), "%d.%m"))] <- "Braunvieh Schweiz and Qualitas AG"
vec_topics_schedule[which(Datum == format(as.Date("2019/05/20"), "%d.%m"))] <- "Questions and Test Exams"
vec_topics_schedule[which(Datum == format(as.Date("2019/05/27"), "%d.%m"))] <- "Final Exams"

# define dataframe for schedule table
df_schedule <- data.frame(Week = Woche, Date = Datum, Topic = vec_topics_schedule, stringsAsFactors = FALSE)
knitr::kable(df_schedule)
```



## Material
The lecture notes, the slides, the exercises and the solutions can be downloaded from this site.

### Course Notes
```{r NotesTable, echo=FALSE, results='asis'}
# url for course notes
s_cn_asm_url <- paste0(sBaseUrl, "cn_asm/")
s_cn_gel_url <- paste0(sBaseUrl, "cn_gel/")

vec_topics_notes <- vec_topics
# add links to the material
vec_topics_notes[which(Datum == format(as.Date("2019/02/18"), "%d.%m"))] <- paste0("[Introduction](", s_cn_asm_url, "asm-intro.pdf", ")") 
vec_topics_notes[which(Datum == format(as.Date("2019/02/25"), "%d.%m"))] <- paste0("[Fixed Linear Effects Model (Part 1)](", s_cn_asm_url, "asm-flem1.pdf", "), [Part2](", s_cn_asm_url, "asm-flem2.pdf)") 
vec_topics_notes[which(Datum == format(as.Date("2019/03/04"), "%d.%m"))] <- paste0("[GBLUP](", s_cn_asm_url, "asm-gblup.pdf)") 
vec_topics_notes[which(Datum == format(as.Date("2019/03/11"), "%d.%m"))] <- paste0("[LASSO](", s_cn_asm_url, "asm-lasso.pdf)") 
vec_topics_notes[which(Datum == format(as.Date("2018/03/18"), "%d.%m"))] <- paste0("[Bayes](", s_cn_asm_url, "asm-bayes.pdf)") 
vec_topics_notes[which(Datum == format(as.Date("2018/03/25"), "%d.%m"))] <- paste0("[Appendix](", s_cn_asm_url, "asm-appendix.pdf)") 

vec_topics_notes[which(Datum == format(as.Date("2019/04/01"), "%d.%m"))] <- paste0("[Introduction](", s_cn_gel_url, "01-intro.pdf)") 
vec_topics_notes[which(Datum == format(as.Date("2019/04/08"), "%d.%m"))] <- paste0("[Breeding Programs](", s_cn_gel_url, "02-bprog.pdf)") 
vec_topics_notes[which(Datum == format(as.Date("2019/04/15"), "%d.%m"))] <- paste0("[Implementation of Breeding Programs](", s_cn_gel_url, "03-implbp.pdf)") 
vec_topics_notes[which(Datum == format(as.Date("2019/04/29"), "%d.%m"))] <- paste0("[Model Selection](", s_cn_gel_url, "04-modsel.pdf)") 
vec_topics_notes[which(Datum == format(as.Date("2019/05/06"), "%d.%m"))] <- paste0("[Variance Components Estimation](", s_cn_gel_url, "05-vce.pdf)") 
vec_topics_notes[which(Datum == format(as.Date("2019/05/13"), "%d.%m"))] <- paste0("[](", s_cn_gel_url, "", ")") 

# define dataframe for schedule table
df_notes <- data.frame(Week = Woche, Date = Datum, Topic = vec_topics_notes, stringsAsFactors = FALSE)
knitr::kable(df_notes)
```


### Slides
```{r SlidesTable, echo=FALSE, results='asis'}
# url for slides
s_sl_url <- paste0(sBaseUrl, "sl/")
vec_topics_slides <- vec_topics
# links zu slides
vec_topics_slides[which(Datum == format(as.Date("2019/02/18"), "%d.%m"))] <- paste0("[Introduction](", s_sl_url, "w01/asm-w01-v01-intro.pdf", ")") 
vec_topics_slides[which(Datum == format(as.Date("2019/02/25"), "%d.%m"))] <- paste0("[Fixed Linear Effects Model](", s_sl_url, "w02/asm-w02-v02-flem.pdf", ")") 
vec_topics_slides[which(Datum == format(as.Date("2019/03/04"), "%d.%m"))] <- paste0("[GBLUP](", s_sl_url, "w03/asm-w03-v03-gblup.pdf)") 
vec_topics_slides[which(Datum == format(as.Date("2019/03/11"), "%d.%m"))] <- paste0("[GRM](", s_sl_url, "w04/asm-w04-v04-grm.pdf), [GBLUP II](", s_sl_url, "w04/asm-w04-v04-gblup.pdf)") 
vec_topics_slides[which(Datum == format(as.Date("2018/03/18"), "%d.%m"))] <- paste0("[LASSO](", s_sl_url, "w05/asm-w05-v05-lasso.pdf)") 
vec_topics_slides[which(Datum == format(as.Date("2018/03/25"), "%d.%m"))] <- paste0("[Bayes](", s_sl_url, "w06/asm-w06-v06-bayes.pdf)") 

vec_topics_slides[which(Datum == format(as.Date("2019/04/01"), "%d.%m"))] <- paste0("[Introduction](", s_sl_url, "w07/gel-w07-v01-intro.pdf)")
vec_topics_slides[which(Datum == format(as.Date("2019/04/08"), "%d.%m"))] <- paste0("[Suisag - Zuchtwertschätzung beim Schwein](", s_sl_url, "w08/Zuchtwertschaetzung_Vorlesung_pst_ETH_Zuerich.pdf)")
vec_topics_slides[which(Datum == format(as.Date("2019/04/15"), "%d.%m"))] <- paste0("[Implementation of Breeding Program](", s_sl_url, "w09/gel-w09-v02-bp.pdf)")
vec_topics_slides[which(Datum == format(as.Date("2019/04/29"), "%d.%m"))] <- paste0("[Model Selection](", s_sl_url, "w11/gel-w11-v03-bp.pdf)")
vec_topics_slides[which(Datum == format(as.Date("2019/05/06"), "%d.%m"))] <- paste0("[Variance Components](", s_sl_url, "w12/gel-w12-v04-vce.pdf)")
vec_topics_slides[which(Datum == format(as.Date("2019/05/13"), "%d.%m"))] <- paste0("[Braunvieh Schweiz](", s_sl_url, "w13/Vortrag_Braunvieh_Schweiz_ETH_Studenten_Mai_2019_mod.pdf), ",
                                                                                    "[Qualitas AG](", s_sl_url, "w13/Vorstellung_Qualitas_20190501.pdf)")
  
# create data frame and convert it to table
df_slides <- data.frame(Week = Woche, Date = Datum, Topic = vec_topics_slides, stringsAsFactors = FALSE)
knitr::kable(df_slides)
```


### Excercises
```{r ExTable, echo=FALSE, results='asis'}
# url for exercises
s_ex_url <- paste0(sBaseUrl, "ex/")
vec_topics_ex <- vec_topics
vec_topics_ex[which(Datum == format(as.Date("2019/02/18"), "%d.%m"))] <- paste0("[](", s_ex_url, "", ")") 
vec_topics_ex[which(Datum == format(as.Date("2019/02/25"), "%d.%m"))] <- paste0("[Breeding Values and Prediction of Effects](", s_ex_url, "w02/asm-w02-ex01.pdf", ")") 
vec_topics_ex[which(Datum == format(as.Date("2019/03/04"), "%d.%m"))] <- paste0("[Fixed Linear Effect Model](", s_ex_url, "w03/asm-w03-ex02.pdf)") 
vec_topics_ex[which(Datum == format(as.Date("2019/03/11"), "%d.%m"))] <- paste0("[Genomic Relationship Matrix](", s_ex_url, "w04/asm-w04-ex03.pdf)") 
vec_topics_ex[which(Datum == format(as.Date("2018/03/18"), "%d.%m"))] <- paste0("[Genomic BLUP](", s_ex_url, "w05/asm-w05-ex04.pdf)") 
vec_topics_ex[which(Datum == format(as.Date("2018/03/25"), "%d.%m"))] <- paste0("[GBLUP II](", s_ex_url, "w06/asm-w06-ex05.pdf)") 
vec_topics_ex[which(Datum == format(as.Date("2019/04/01"), "%d.%m"))] <- paste0("[Bayes](", s_ex_url, "w07/asm-w07-ex06.pdf)")
vec_topics_ex[which(Datum == format(as.Date("2019/04/08"), "%d.%m"))] <- paste0("[](", s_ex_url, "", ")")
vec_topics_ex[which(Datum == format(as.Date("2019/04/15"), "%d.%m"))] <- paste0("[Model Selection](", s_ex_url, "w09/gel-w09-ex01.pdf)")
vec_topics_ex[which(Datum == format(as.Date("2019/04/29"), "%d.%m"))] <- paste0("[](", s_ex_url, "", ")")
vec_topics_ex[which(Datum == format(as.Date("2019/05/06"), "%d.%m"))] <- paste0("[Variance Components Estimation](", s_ex_url, "w12/gel-w12-ex02.pdf)")
vec_topics_ex[which(Datum == format(as.Date("2019/05/13"), "%d.%m"))] <- paste0("[Prediction of Breeding Values](", s_ex_url, "w13/gel-w13-ex03.pdf)")

# create dataframe and table
df_ex <- data.frame(Week = Woche, Date = Datum, Topic = vec_topics_ex, stringsAsFactors = FALSE)
knitr::kable(df_ex)
```



### Solutions
```{r SolTable, echo=FALSE, results='asis'}
s_sol_url <- paste0(sBaseUrl, "sol/")
vec_topics_sol <- vec_topics
vec_topics_sol[which(Datum == format(as.Date("2019/02/18"), "%d.%m"))] <- paste0("[](", s_sol_url, "", ")") 
vec_topics_sol[which(Datum == format(as.Date("2019/02/25"), "%d.%m"))] <- paste0("[Breeding Values and Prediction of Effects](", s_sol_url, "w02/asm-w02-sol01.pdf", ")") 
vec_topics_sol[which(Datum == format(as.Date("2019/03/04"), "%d.%m"))] <- paste0("[Fixed Linear Effect Model](", s_sol_url, "w03/asm-w03-sol02.pdf)")
vec_topics_sol[which(Datum == format(as.Date("2019/03/11"), "%d.%m"))] <- paste0("[Genomic Relationship Matrix](", s_sol_url, "w04/asm-w04-sol03.pdf)")
vec_topics_sol[which(Datum == format(as.Date("2018/03/18"), "%d.%m"))] <- paste0("[Genomic BLUP](", s_sol_url, "w05/asm-w05-sol04.pdf)")
vec_topics_sol[which(Datum == format(as.Date("2018/03/25"), "%d.%m"))] <- paste0("[GBLUP II](", s_sol_url, "w06/asm-w06-sol05.pdf)")
vec_topics_sol[which(Datum == format(as.Date("2019/04/01"), "%d.%m"))] <- paste0("[Bayes](", s_sol_url, "w07/asm-w07-sol06.pdf)")
vec_topics_sol[which(Datum == format(as.Date("2019/04/08"), "%d.%m"))] <- paste0("[](", s_sol_url, "", ")")
vec_topics_sol[which(Datum == format(as.Date("2019/04/15"), "%d.%m"))] <- paste0("[](", s_sol_url, "", ")")
vec_topics_sol[which(Datum == format(as.Date("2019/04/29"), "%d.%m"))] <- paste0("[Model Selection](", s_sol_url, "w11/gel-w09-sol01.pdf)")
vec_topics_sol[which(Datum == format(as.Date("2019/05/06"), "%d.%m"))] <- paste0("[Variance Components Estimation](", s_sol_url, "w12/gel-w12-sol02.pdf)")
# vec_topics_sol[which(Datum == format(as.Date("2019/05/13"), "%d.%m"))] <- paste0("[Prediction of Breeding Values](", s_sol_url, "w13/gel-w13-sol03.pdf)")

# create dataframe and table
df_sol <- data.frame(Week = Woche, Date = Datum, Topic = vec_topics_sol, stringsAsFactors = FALSE)
knitr::kable(df_sol)
```



### Additional Material
```{r AdditionalMaterial, echo=FALSE, results='asis'}
s_misc_url <- paste0(sBaseUrl, "misc/")
vec_topics_addmat <- vec_topics
vec_topics_addmat[which(Datum == format(as.Date("2019/02/25"), "%d.%m"))] <- paste0("[Genotype Data Exercise 1](", s_ex_url, "w02/asm_w02_ex01_p02_genodatafile.csv", ")") 
vec_topics_addmat[which(Datum == format(as.Date("2019/03/18"), "%d.%m"))] <- paste0("[Data Analysis Protocol](", s_misc_url, "20190318_demo_sol4p01/20190318_demo_sol4p01.html)")
vec_topics_addmat[which(Datum == format(as.Date("2019/04/15"), "%d.%m"))] <- paste0("[Dataset Model Selection](", s_ex_url, "w09/data_bp_w09.csv)")
vec_topics_addmat[which(Datum == format(as.Date("2019/05/06"), "%d.%m"))] <- paste0("[Dataset Variance Components](", s_ex_url, "w12/data_bp_w12.csv), ",
                                                                                    "[Dataset Sire Model](", s_ex_url, "w12/data_sire_w12.csv)")

# dataframe and table with additional material
df_addmat <- data.frame(Week = Woche, Date = Datum, Topic = vec_topics_addmat, stringsAsFactors = FALSE)
knitr::kable(df_addmat)
```


```{r, echo=FALSE, results='asis'}
cat(rmddochelper::get_latest_change(ps_msg = "Latest Changes"), "\n")
```
